---
title: "OznalProject"
output: html_document
---

# Predicting Travel Time of Bratislava MHD

## Selecting Data

For the project we selected a publicly available local open dataset available here:

<https://data.bratislava.sk/datasets/aba12fd2cbac4843bc7406151bc66106/about>

This dataset contains raw data collected by MHD Bratislava which are stored in the form of multiple text files connected by foreign keys (similar to relational databases).

The newest data update our project use was released at:

April 17, 2026

## Defining Hypothesis

The project is hypothesis-driven meaning a hypothesis will be evaluated at its end. The hypothesis is stated as follows:

*"We hypothesize that scheduled travel time between two stops in GTFS systems can be predicted from such metrics like euclidean distance, route & stop attributes, geographical metrics like Lon and Lat points, temporal features like arrival times and other metrics."*

## Selecting Scenarios

Based on the hypothesis and data structure we decided to go with these two scenarios:

1.  (Scenario 1) Compare a regression or classification setup across three statistical or machine-learning methods and two feature-space partitioning approaches. Identify which features remain informative across both method families, discuss the implications for explainability, and outline the benefits each approach offers.
2.  (Scenario 3) Extend your work by comparing algorithmic and embedded feature-selection methods for the same regression or classification task. This task may require some research on your side, as not all necessary information will be taught through the course. Report how many features you ultimately retain, and which features lose statistical significance as you add or remove predictors.

## Analyzing Data

As we mentioned earlier, the we data represent GTFS system of MHD Bratislava which is stored in several .txt files such as:

1.  *stops.txt* (attributes about stops located in Bratislava like Hlavná stanica, Pod stanicou, etc.)
2.  *trips.txt* (a concrete trip taken by concrete vehicle following concrete stops and of conrete route)
3.  *stop_times.txt* (attributes about stop times of concrete trip at concrete stop)
4.  routes.txt (attributes about routes such as route_type managed by MHD)
5.  and other files

Many of the files or columns are highly sparse and are not provided by MHD (e.g. *frequencies.txt*, *pathways.txt* or columns *bike_allowed*, *stop_headsign* or *shape_dist_traveled).*

When analyzing the dimensions of data it has almost **700k rows** and almost **70** which should be more than satisfactory for training simple regression models like *Linear Regression* or even more complex tree-based models like *Random Forests*.

## EDA

### Libraries

```{r}
library("tidyverse")
library("readr")
library("dplyr")
```

### Loading Data

Apart from loading dataset, we also perform **stratified_sampling** where we pick 10% of observations for each day_type. This ensures that each category is represented with respect to its proportion in data.

```{r}

timetable <- read_csv("../data/interim/dataset.csv")

table(timetable$day_type)

timetable <- timetable %>%
  group_by(day_type) %>%  # important variable
  sample_frac(0.1) %>%   # 10% sample
  ungroup()
```

Now we can take a lot at the dataset to see its structure and info about its columns.

```{r}

glimpse(timetable)
```

### Data Preparation

By performing brief initial inspection of data we noticed several problems:

1.  **Missing or sparse** (some columns clearly contain NA values)
2.  **Constant or Near-Constant Columns**
3.  **Uninformative columns** (urls, descriptions, etc.)
4.  **Unsupported columns types** (strings, time, etc.)

These problems and more will need to be dealt with during this phase. We are going to turn this unsuitable format into nice training-ready data.

#### Utility Functions

In this section we define a set of utility function that we are going to use in the later phases. Both of the functions do the same - they take a vector of columns names or row indices and drop it from the dataset. Apart from this, they also print some nice logs into console about the reduction.

```{r}
drop_cols_report <- function(data, cols_to_drop) {
  # 1. Store the initial count
  col_num_before <- ncol(data)
  
  # 2. Perform the drop
  # We use any_of() instead of all_of() to prevent errors 
  # if a column was already dropped previously.
  data_reduced <- data %>%
    select(!any_of(cols_to_drop))
  
  # 3. Calculate difference
  col_num_after <- ncol(data_reduced)
  dropped_count <- col_num_before - col_num_after
  
  # 4. Print the report
  message(paste0("Dropped cols: ", dropped_count, 
                 " - Current size: ", col_num_after))
  
  # 5. Return the cleaned dataframe
  return(data_reduced)
}

drop_rows_report <- function(data, rows_to_drop) {
  # 1. Store initial count
  row_num_before <- nrow(data)
  
  # 2. Perform the drop
  # Using -rows_to_drop handles the removal by index.
  # We check if rows_to_drop is empty first to avoid returning an empty table.
  if (length(rows_to_drop) > 0) {
    data_reduced <- data[-rows_to_drop, ]
  } else {
    data_reduced <- data
  }
  
  # 3. Calculate difference
  row_num_after <- nrow(data_reduced)
  dropped_count <- row_num_before - row_num_after
  
  # 4. Print the report
  message(paste0("Dropped rows: ", dropped_count, 
                 " - Current size: ", row_num_after))
  
  return(data_reduced)
}

find_duplicate_columns <- function(df) {
  # dup_col_indexes <- which(duplicated(as.list(df)))
  # dup_cols <- names(df)[dup_col_indexes]
  # 
  # return(dup_cols)
  is_dup <- duplicated(as.list(df))
  
  dup_summary <- tibble::tibble(
    column = names(df),
    is_duplicate = is_dup
  )
  
  return(dup_summary %>% filter(is_duplicate) %>% pull(column))
}

# =========================================================
# 1. One-hot encoding for low-cardinality predictors
#    Returns the original dataset with encoded columns added
#    and the original column removed.
# =========================================================
# one_hot_encode <- function(data, col) {
#   data[[col]] <- as.factor(data[[col]])
#   
#   mm <- model.matrix(
#     as.formula(paste0("~ `", col, "` - 1")),
#     data = data
#   ) %>%
#     as.data.frame()
#   
#   data %>%
#     bind_cols(mm) %>%
#     select(-all_of(col))
# }
one_hot_encode <- function(data, col) {
  data[[col]] <- as.factor(data[[col]])
  
  mm <- model.matrix(
    as.formula(paste0("~ `", col, "`")),
    data = data
  ) %>%
    as.data.frame()
  
  # remove intercept column
  mm <- mm %>%
    dplyr::select(-`(Intercept)`)
  
  data %>%
    dplyr::bind_cols(mm) %>%
    dplyr::select(-all_of(col))
}

# =========================================================
# 2. Frequency encoding
#    Replaces each category by its relative frequency.
#    Good for medium/high-cardinality categoricals.
# =========================================================
frequency_encode <- function(data, col, new_col = NULL) {
  if (is.null(new_col)) {
    new_col <- paste0(col, "_freq")
  }
  
  freq_tbl <- data %>%
    count(.data[[col]], name = "n") %>%
    mutate(freq = n / sum(n)) %>%
    select(all_of(col), freq)
  
  data %>%
    left_join(freq_tbl, by = col) %>%
    rename(!!new_col := freq)
}

# =========================================================
# 3. Hex color -> RGB
#    Good for route_color if you want numeric representation
#    without category explosion.
# =========================================================
hex_to_rgb_df <- function(data, col = "route_color", prefix = "route") {
  hex_to_rgb <- function(hex) {
    hex <- gsub("#", "", as.character(hex))
    if (is.na(hex) || nchar(hex) != 6) return(c(NA, NA, NA))
    c(
      strtoi(substr(hex, 1, 2), 16L),
      strtoi(substr(hex, 3, 4), 16L),
      strtoi(substr(hex, 5, 6), 16L)
    )
  }
  
  rgb_vals <- t(sapply(data[[col]], hex_to_rgb)) %>%
    as.data.frame()
  
  colnames(rgb_vals) <- c(
    paste0(prefix, "_r"),
    paste0(prefix, "_g"),
    paste0(prefix, "_b")
  )
  
  bind_cols(data, rgb_vals)
}
```

#### Handling Missing Values

Throughout the following sections we are going to do the same:

1.  Create summaries about columns or rows which are somehow problematic
2.  Extract the column names or row indexes in the form of vectors

This structured format ensures that they can be effectively dropped by the reusable functions defined earlier.

In the below code we create a summary of missing values for every columns using *is.na()* provided by base R. The summaries are then split into 2 vectors containing:

1.  **low-missing features** (0\>missing_pct\>50): can be kept and handled by methods like imputing, etc.
2.  **high-missing features**: (missing_pct\>=50): problematic - should be dropped right away

```{r}
# 1. Missing values
# =========================

missing_summary <- timetable %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "missing_count") %>%
  mutate(missing_pct = missing_count / nrow(timetable) * 100) %>%
  arrange(desc(missing_pct))

missing_summary

# ===
# 1.1 Low_missing features (can be kept - must be imputed)
# ===

low_missing <- missing_summary %>%
  filter(missing_pct > 0 & missing_pct < 50) %>%
  pull(column)

low_missing

# ===
# 1.2 High missing features (must be removed)
# ===

high_missing <- missing_summary %>%
  filter(missing_pct >= 50) %>%
  pull(column)

high_missing

```

#### Constant & Near-Constant Features

These are features that carry no or only minimal signal and would only cause model to overfit they should be dropped too.\
We identify near-constant features by calculating the ratio between the 2 largest frequencies identified in a column. If this ratio is above 20 we consider this column to be near-constant.

```{r}
# =========================
# 2. Constant & Near-Constant features
# =========================
variance_summary <- timetable %>%
  summarise(across(everything(), ~ n_distinct(.))) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "n_unique") %>%
  arrange(desc(n_unique))
variance_summary

# ===
# 2.1 Constant features (should be dropped immediately)
# ===

constant_features <- variance_summary %>%
  filter(n_unique == 1) %>% pull(column)

constant_features

# ===
# 2.2 Near-zero variance (should be dropped too, too weak predictors)
# ===


# formula: ratio = freq_1 / freq_2
nzv_summary <- timetable %>%
  summarise(across(everything(), ~ {
    freq <- sort(table(.), decreasing = TRUE)
    if (length(freq) <= 1) return(NA)
    freq[1] / freq[2]
  })) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "freq_ratio") %>%
  arrange(desc(freq_ratio))

print(nzv_summary)

# Threshold example: freq_ratio > 20
nzv_cols <- nzv_summary %>%
  filter(freq_ratio > 20) %>%
  pull(column)

print(nzv_cols)
```

#### Handling Duplicates

Now we also verify whether dataset doesn't contain duplicate rows (stored as vector of row indices) and duplicate columns (stored as vector of column names). Please note that function *find_duplicate_cols* iteratively examines column by column comparing it with previous ones. This way it also finds such columns which contain only NaN values.

```{r}

# =========================
# 4. Handling duplicates
# =========================

# ===
# 4.1 Duplicate rows (should be removed)
# ===

dup_row_indexes <- timetable %>%
  duplicated() %>%
  which()

print(paste("Duplicate rows:", dup_row_indexes))

# ===
# 4.2 Duplicate cols (should be removed)
# ===


# More efficient check for duplicate columns
# R looks at the "top level" of the data frame. duplicated() now looks at each 
# vector (column) as a single item in a list and compares them.
# dup_col_indexes <- which(duplicated(as.list(timetable)))
# dup_cols <- names(timetable)[dup_col_indexes]
dup_cols <- find_duplicate_columns(timetable)

paste("Duplicate cols")
dup_cols
```

#### High-cardinality Columns

In this section we create summary about cardinality of character columns by using *n_distinct* and *n_unique* properties. These columns will need to be handled differently from one-hot-encoding since this would result in very large dataset.

```{r}

# =========================
# 5. High cardinality categorical variables
# =========================
cat_summary <- timetable %>%
  select(where(is.character)) %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "n_unique") %>%
  arrange(desc(n_unique))

print(cat_summary)

# ===
# Columns with too many categories (characters)
# ===
high_cardinality <- cat_summary %>%
  filter(n_unique > 10)
high_cardinality

high_cardinality_cols <- high_cardinality %>% pull(column)
high_cardinality_cols
```

#### Feature Reduction

The problematic features like highly-sparse features, constant or near-constant features are quite useless for models and should be dropped right away using the utility functions.

```{r}
timetable_reduced <- drop_cols_report(timetable, high_missing)
timetable_reduced <- drop_cols_report(timetable_reduced, constant_features)
timetable_reduced <- drop_cols_report(timetable_reduced, nzv_cols)
timetable_reduced <- drop_rows_report(timetable_reduced, dup_row_indexes)
timetable_reduced <- drop_cols_report(timetable_reduced, dup_cols)

dup_cols
glimpse(timetable_reduced)

table(timetable_reduced$arrival_time_second)
# # ===
# # 7.7 Highly-correlated cols
# # ===
# 
# timetable_reduced <- drop_cols_report(timetable_reduced, highly_correlated)

# glimpse(timetable_reduced)

# ===
# 7.8 Irrelevant cols
# ===
# these were used during extraction process and were broken into several predictors
# ===


highly_correlated_char = c('arrival_time', 'service_id', 'shape_id', 'stop_id',
                           'monday', 'tuesday', 'wednesday', 'thursday', 'friday',
                           'saturday', 'sunday', 'trip_id', 'stop_sequence')
timetable_reduced <- drop_cols_report(timetable_reduced, highly_correlated_char)

# ===
# 7.8 Irrelevant cols
# ===

# these should not be in the dataset after extraction update and remove from here!!
# irrelevant_features = c('day_type')

# timetable_reduced <- drop_cols_report(timetable_reduced, irrelevant_features)

# ===
# 7.9 Basic info after filtering
# ===
dim(timetable_reduced)
glimpse(timetable_reduced)

```

#### Handling low-missing features

Let's first analyze the low-missing features:

```{r}
missing_summary %>% filter(missing_pct < 50 & missing_pct > 0)
```

##### time_since_last_stop

This one is calculated as:

*time_since_last_stop = arrival_seconds - prev(departure_seconds)*

where *prev()* represents the previous stop within a trip (sorted by *stop_sequence*). The NaN values logically represent the first stop within a trip which doesn't have a previous stop. Thus, we apply imputation with 0:

```{r}
summary(timetable_reduced$time_since_last_stop)
```

```{r}
# ===
# 8.1 Time_since_last_stop
# ===
# These are first stops in a trip

timetable_imputed <- timetable_reduced %>%
  mutate(
    time_since_last_stop = replace_na(time_since_last_stop, 0)
  )
```

##### wheelchair_accessible

Since wheelchair_accessible mostly contains value 1 (which means accessible) and a very small portion of missing values it cannot serve as a reliable predictor and should be dropped too.

```{r}
table(timetable_imputed$wheelchair_accessible)
timetable_imputed <- drop_cols_report(timetable_imputed, c('wheelchair_accessible'))
```

```{r}
# =========================
# 9. Handling categorical values
# =========================

# # Scan all character columns and count unique values
# unique_counts <- timetable_imputed %>%
#   summarise(across(where(is.character), ~n_distinct(.))) %>%
#   pivot_longer(cols = everything(), names_to = "column_name", values_to = "unique_count")
# 
# print(unique_counts)
# print(cat_summary)
```

#### Feature Encoding

Now, the remaining preprocessing step focuses on encoding categorical and non-ordinal features. Three main techniques were applied:

1.  **One-hot encoding**: Used for categorical features with relatively low cardinality (typically fewer than 15 unique values) to preserve interpretability while avoiding excessive dimensionality.
2.  **Frequency encoding**: Applied to high-cardinality categorical features, where one-hot encoding would lead to a large number of sparse columns. This approach replaces each category with its relative frequency, retaining distributional information in a compact form.
3.  **Feature transformation and decomposition**: Certain features were transformed into more informative numerical representations. For example, hexadecimal color codes were decomposed into RGB components.

```{r}
glimpse(timetable_imputed)

timetable_encoded <- timetable_imputed %>%
  one_hot_encode("platform_code") %>%
  one_hot_encode("day_type") %>%
  one_hot_encode("zone_id") %>%
  one_hot_encode("direction_id") %>%
  # one_hot_encode("wheelchair_accessible") %>%
  one_hot_encode("pickup_type") %>%
  one_hot_encode("route_type") %>%
  hex_to_rgb_df("route_color", prefix = "route") %>%
  frequency_encode("route_short_name") %>%
  frequency_encode("route_id") %>%
  frequency_encode("trip_headsign") %>%
  frequency_encode("stop_name") %>%
  frequency_encode("stop_code") %>%
  select(
    -route_color,
    -route_short_name,
    -route_id,
    -trip_headsign,
    -stop_name,
    -stop_code
  )

dup_cols_2 <- find_duplicate_columns(timetable_encoded)
dup_cols_2

# timetable_encoded$daytimetable_encoded
glimpse(timetable_encoded)
```
#### Detecting duplicates
Let's check again whether there are no duplicate columns in the dataset.

```{r}
dup_cols_2 <- find_duplicate_columns(timetable_encoded)
dup_cols_2

timetable_encoded <- drop_cols_report(timetable_encoded, dup_cols_2)
```


#### Saving Dataset

```{r}
# =========================
# 10. Save dataset
# =========================
saveRDS(timetable_encoded, "./datasets/timetable_encoded.rds")
```

### Training Preparation

Now that issues such as missing values, constant or near-constant features, and appropriate encoding have been addressed, the resulting dataset represents a set of candidate predictors suitable for model training.

In this second phase of exploratory data analysis, the focus shifts from general data quality to model-specific requirements. In particular, we examine:

1. **Multicollinearity** among predictors, which can negatively affect linear models
2. **Extreme outliers** in feature distributions, which may influence model stability
3. **The nature of relationships** between predictors and the target variable, including potential non-linear patterns

To support this analysis, various visualization techniques such as heatmaps, boxplots, and scatterplots are employed using libraries like ggplot2. These tools help to better understand how the data aligns with the assumptions and behavior of different regression models.

Based on these insights, appropriate transformations and feature selection strategies are proposed to improve model performance and robustness.

#### Libraries

```{r}
library("ggplot2")
library("patchwork")
```

#### Loading Data

```{r}

timetable_encoded <- readRDS("./datasets/timetable_encoded.rds")
glimpse(timetable_encoded)

```
#### Utility Functions

This section contains utility functions that are going to be repeteadly used throughout this phase.

```{r}

select_correlated_features <- function(
    data,
    target_col,
    min_abs_cor = 0.1,
    top_n = NULL,
    always_keep = NULL
) {
  # keep only numeric columns for correlation
  numeric_data <- data %>%
    dplyr::select(where(is.numeric))
  
  # check target exists
  if (!(target_col %in% names(numeric_data))) {
    stop(paste("Target column", target_col, "must exist and be numeric."))
  }
  
  # compute correlations with target
  cor_vec <- cor(numeric_data, use = "complete.obs")[, target_col]
  
  cor_df <- data.frame(
    feature = names(cor_vec),
    correlation = as.numeric(cor_vec)
  ) %>%
    dplyr::filter(feature != target_col) %>%
    dplyr::mutate(abs_correlation = abs(correlation)) %>%
    dplyr::arrange(desc(abs_correlation))
  
  # threshold filter
  cor_df <- cor_df %>%
    dplyr::filter(abs_correlation >= min_abs_cor)
  
  # optional top_n restriction
  if (!is.null(top_n)) {
    cor_df <- cor_df %>%
      dplyr::slice_head(n = top_n)
  }
  
  # final feature list
  selected_features <- unique(c(
    target_col,
    cor_df$feature,
    always_keep
  ))
  
  # keep only existing columns (safe)
  selected_features <- intersect(selected_features, names(data))
  
  return(data %>%
           dplyr::select(all_of(selected_features)))
}

plot_correlation_heatmap <- function(df, threshold = 0.3, title = "Correlation Heatmap") {
  
  # 1. Calculate correlation matrix (only for numeric columns)
  # It's safer to filter for numeric variables automatically
  cor_matrix <- df %>%
    select(where(is.numeric)) %>%
    cor(use = "complete.obs")
  
  # 2. Convert to long format
  cor_long <- as.data.frame(as.table(cor_matrix))
  colnames(cor_long) <- c("Var1", "Var2", "Correlation")
  
  # 3. Create the plot
  ggplot(cor_long, aes(x = Var1, y = Var2, fill = Correlation)) +
    geom_tile() +
    # Dynamic label filtering based on the 'threshold' argument
    geom_text(
      aes(label = ifelse(abs(Correlation) > threshold, round(Correlation, 2), "")),
      size = 3,
      color = "black"
    ) +
    scale_fill_gradient2(
      low = "#377eb8",  # Nice blue
      mid = "white",
      high = "#e41a1c", # Nice red
      midpoint = 0,
      limit = c(-1, 1)   # Ensures the scale is always centered
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.title = element_blank(),
      panel.grid = element_blank()
    ) +
    ggtitle(title)
}

plot_box_dist <- function(df, var_name) {
  ggplot(df, aes(x = !!sym(var_name))) +
    # Adding a dummy y-axis value or a factor(0) helps boxplot orientation
    geom_boxplot(fill = "lightblue", outlier.color = "red", outlier.alpha = 0.5) +
    coord_flip() +
    theme_minimal() +
    labs(
      title = paste(var_name),
      x = NULL # Removing x label as the title covers it
    )
}

plot_predictor_boxplots <- function(
    df,
    target_col = "segment_time_s",
    ncol = 3,
    show_message = TRUE,
    title = "Outlier Detection Across Predictors",
    subtitle = "Boxplots identifying potential outliers in red"
) {
  # Identify binary columns
  is_binary <- purrr::map_lgl(df, ~ length(unique(stats::na.omit(.x))) <= 2)
  binary_cols <- names(is_binary[is_binary])
  
  # Select predictors to plot
  predictors_to_plot <- setdiff(names(df), c(target_col, binary_cols))
  
  # Optional message
  if (show_message) {
    message("Dropping binary features: ", paste(binary_cols, collapse = ", "))
  }
  
  # Create individual plots
  plot_list <- purrr::map(predictors_to_plot, ~ plot_box_dist(df, .x))
  
  # Combine plots
  combined_plot <- patchwork::wrap_plots(plot_list, ncol = ncol) +
    patchwork::plot_annotation(
      title = title,
      subtitle = subtitle,
      theme = ggplot2::theme(
        plot.title = ggplot2::element_text(size = 18, face = "bold")
      )
    )
  
  return(combined_plot)
}

```

#### Data Engineering

In order to better capture temporal patterns present in the data, the original time variable was transformed into a cyclical representation. Specifically, the variable seconds_since_midnight was mapped onto the unit circle using sine and cosine transformations. This approach preserves the periodic nature of time, ensuring that values close in real-world time (e.g., late night and early morning) are also close in the transformed feature space.

In addition to temporal encoding, interaction features were introduced to model the combined effect of distance and time of day. The features distance_sin_interaction and distance_cos_interaction were created by multiplying the segment distance with the corresponding sine and cosine components. These interaction terms allow the model to capture time-dependent variations in travel behavior, such as differing traffic conditions across the day.

```{r}
timetable_engineered <- timetable_encoded %>%
  mutate(
    sin_time = sin(2 * pi * seconds_since_midnight / 86400),
    cos_time = cos(2 * pi * seconds_since_midnight / 86400),
    distance_sin_interaction = segment_distance_m * sin_time,
    distance_cos_interaction = segment_distance_m * cos_time,
  )
```

##### Visualizations

Now let's visualize the engineered features' distributions by plotting histograms:

```{r}
timetable_engineered %>%
  select(sin_time, cos_time, distance_sin_interaction, distance_cos_interaction) %>%
  pivot_longer(cols = everything(), names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  facet_wrap(~ feature, scales = "free") +
  theme_minimal()
```

The distributions of the cyclical features *sin_time* and *cos_time* exhibit smooth, symmetric patterns consistent with their trigonometric formulation. Both variables show higher densities near their extreme values (−1 and 1), with a gradual decline toward intermediate values, followed by a mirrored increase.


The interaction features, *distance_sin_interaction* and *distance_cos_interaction*, are centered around zero and exhibit approximately bell-shaped distributions, suggesting a near-normal structure. However, the presence of heavy tails reflects the influence of large distance values, resulting in more extreme interaction effects.

To verify the distributions of *sin_time* and *cos_time* even further we can do **circular structure check**.

```{r}
ggplot(timetable_engineered, aes(x = sin_time, y = cos_time)) +
  geom_point(alpha = 0.1) +
  coord_fixed() +
  theme_minimal()
```

The joint visualization of these features confirms their correct transformation, forming a well-defined circular structure, which indicates that the periodic nature of time has been successfully preserved.



```{r}

subset_data <- select_correlated_features(
  data = timetable_encoded,
  target_col = "segment_time_s",
  min_abs_cor = 0.05,
  always_keep = c("seconds_since_midnight", "route_short_name_freq")
)

glimpse(subset_data)


# subset_data <- timetable_encoded %>%
#   select(all_of(append('segment_time_s', selected_cor_df %>% pull(feature))))

# ===
# 10.1 Remove multicollinearity
# ===

# collinear_features <- c('route_short_name139', 'trip_part_1139009',
# 'route_colorF56200', 'trip_headsignCintorín Slávičie')
collinear_features <- c()

subset_data_filtered <- subset_data %>%
  select(-all_of(collinear_features))

```

#### Checking multicollinearity & target collinearity

One of the requirements of regression models like linear or polynomial regression is that its predictors should not be heavily correlated. To verify this in a neat way we are going to print the correlation matrix in the form of **HeatMap**.

Since there is a lot of features currently in the dataset we apply this 

```{r}


timetable_selected <- select_correlated_features(timetable_engineered, min_abs_cor = 0.05, target_col = "segment_time_s")

glimpse(timetable_selected)

# summary(timetable_engineered$arrival_seconds)


plot_correlation_heatmap(timetable_selected, threshold = 0.01)
```
From the observed results we can tell that there is no serious collinearity among features apart from *cos_time* and *distance_cos_interaction*

```{r}
correlated_features <- c('cos_time', 'distance_cos_interaction', 'sin_time', 'distance_sin_interaction',
                         'arrival_seconds', 'seconds_since_midnight')

plot_correlation_heatmap(
  timetable_engineered %>% 
    select(all_of(c(correlated_features, "segment_time_s")))
)

```

```{r}
highly_collineared_features <- c('zone_id101')
timetable_engineered <- drop_cols_report(timetable_engineered, highly_collineared_features)
```
#### Outliers

Extreme outliers are also problematic for regression models. Let's visualize them using boxplots.

##### Boxplots

```{r}
combined_plot <- plot_predictor_boxplots(timetable_selected)
print(combined_plot)
```
From the bopxlots that mainly distance and time feature and their interactions have a lot of outliers and that heir distributions are hard to see. Let's analyze this further by inspecting their distributions

##### Distributions

```{r}
problematic_features <- c('segment_distance_m', 'distance_cos_interaction',
                          'distance_sin_interaction', 'time_since_last_stop')

plot_boxplot <- function(data, feature){
  ggplot(data, aes(x = .data[[feature]])) +
    geom_histogram(bins = 50, fill = "steelblue") +
    # scale_x_log10() +
    theme_minimal()
}

plot_boxplot(timetable_selected, 'segment_distance_m')
plot_boxplot(timetable_selected, 'time_since_last_stop')

summary(timetable_selected %>% select(all_of(problematic_features)))

```
##### Transformations

Based on the analysis of the predictors with most outliers and their distributions, we recommend following before using them in training:

1. *segment_distance_m* -> very long tail, strong right skew, **log_transform** recommended
2. *distance_interactions* -> also heavy tails, caused by distance metric. Log cannot be applied for negative values, so transforming distance and the recomputing is recommended
3 *time_since_last_stop* -> only moderate skew and some extremes, recommended removal if not significant model

### Saving Data

```{r}
saveRDS(timetable_engineered, './datasets/timetable_engineered.rds')
```



## Model Training

### Libs

```{r}
library("purrr")
library("car")
```

### Loading Data

```{r}
timetable <- readRDS("./datasets/timetable_engineered.rds")

```

### Transformations

```{r}
# plot_predictor_boxplots(train_data, target_col = 'segment_time_s')

# summary(train_data %>% select(all_of(c('stop_lat', 'arrival_time_second', 'segment_distance_m'))))
timetable_transformed <- timetable_engineered %>%
  mutate(
    segment_distance_m = log1p(segment_distance_m)
  )

plot_predictor_boxplots(timetable_transformed %>% select(c('segment_distance_m')), target_col = 'segment_time_s')

```
```{r}
sample_data <- timetable_transformed %>%
  slice_sample(n = 5000)

train_col_names <- colnames(sample_data)

train_col_names <- train_col_names[!train_col_names %in% c('segment_time_s')]

# length(train_col_names)
length(train_col_names)

# train_col_names <- train_col_names[]

train_col_names <- c("segment_distance_m", 'distance_cos_interaction', 'time_since_last_stop', 'distance_sin_interaction',
                     'stop_lat', 'stop_lon', 'sin_time', 'cos_time')


scatterplot_list <- map(train_col_names, ~plot_scatterplot_line(sample_data, .x))

combined_plot <- patchwork::wrap_plots(scatterplot_list, ncol = 2)
combined_plot
```


### Train/Test Split

```{r}
# ===
# 12.1 Train / test split
# ===

set.seed(42)

# 80/20 split
train_idx <- sample(seq_len(nrow(timetable_transformed)), size = 0.8 * nrow(timetable))

train_data <- timetable[train_idx, ]
test_data  <- timetable[-train_idx, ]


```

### Linear Regression

#### Feature Reduction


```{r}
# unimportant_features <- c('time_since_last_stop', 'platform_codeH', 'arrival_seconds')
# unimportant_features <- c('route_b', 'arrival_seconds', 'route_short_name_freq')

unimportant_features <- c('seconds_since_midnight', 'platform_codeD',
                          'platform_codeG', 'day_typesaturday', 'day_typesunday',
                          'platform_codeF',
                          'platform_codeB', 'platform_codeC',
                          'direction_id1', 'platform_codeJ', 'platform_codeI', 'route_type11',
                          'arrival_seconds', 'arrival_time_hour')

# vif>5, handles collinearity
high_vif <- c('arrival_time_hour', 'arrival_seconds',
              'pickup_type3', 'route_type11')

high_vif <- c()

train_data <- train_data %>%
  select(-all_of(append(unimportant_features, high_vif)))

test_data <- test_data %>%
  select(-all_of(append(unimportant_features, high_vif)))

plot_correlation_heatmap(train_data, threshold = 0.1)
```

#### Linearity

# Create individual plots
  plot_list <- purrr::map(predictors_to_plot, ~ plot_box_dist(df, .x))
  
  # Combine plots
  combined_plot <- patchwork::wrap_plots(plot_list, ncol = ncol) +

```{r}
### Linearity

plot_scatterplot_line <- function(data, x, y = 'segment_time_s'){
  ggplot(data, aes(x = .data[[x]], y = .data[[y]])) +
    geom_point(alpha = 0.2) +
    geom_smooth(method = "lm", color = "red", se = FALSE) +
    geom_smooth(method = "loess", color = "blue", se = FALSE) +
    theme_minimal()
}

sample_data <- train_data %>%
  slice_sample(n = 5000)

train_col_names <- colnames(sample_data)

train_col_names <- train_col_names[!train_col_names %in% c('segment_time_s')]

# length(train_col_names)

# train_col_names <- train_col_names[30:]


scatterplot_list <- map(train_col_names, ~plot_scatterplot_line(sample_data, .x))

combined_plot <- patchwork::wrap_plots(scatterplot_list, ncol = 2)
combined_plot
                                   `
```

#### Model Fit

```{r}
# summary(train_data$seconds_since_midnight)
# lm_model <- lm(segment_time_s ~ ., data = train_data)
lm_poly <- lm(
  segment_time_s ~ 
    segment_distance_m +
    # distance_sin_interaction +
    I(distance_sin_interaction^2) +
    # distance_cos_interaction +
    I(distance_cos_interaction^2) + .,
  data = train_data
)
summary(lm_poly)

vif(lm_poly)
```
#### Homoscedasticity

```{r}
plot(lm_model$fitted.values, lm_model$residuals)
abline(h = 0, col = "red")

data.frame(
  fitted = fitted(lm_model),
  residuals = residuals(lm_model)
) %>%
  ggplot(aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal()
```
#### Evaluation

```{r}
predictions_lm <- predict(lm_poly, newdata = test_data)
```

##### Errors

```{r}
rmse <- sqrt(mean((test_data$segment_time_s - predictions_lm)^2))
rmse
```

```{r}
mae <- mean(abs(test_data$segment_time_s - predictions_lm))
mae
```

```{r}
ss_total <- sum((test_data$segment_time_s - mean(test_data$segment_time_s))^2)
ss_res   <- sum((test_data$segment_time_s - predictions_lm)^2)
r2_test <- 1 - (ss_res / ss_total)
r2_test
```

##### Plots

###### Predicted Vs. Actual

```{r}

plot_predicted_vs_actual <- function(data, predictions, target = 'segment_time_s'){

ggplot(data.frame(actual = data[[target]], pred = predictions),
       aes(x = actual, y = pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_minimal() +
  ggtitle("Predicted vs Actual")
}

plot_predicted_vs_actual(test_data, predictions_lm)

```


###### Residuals vs. Predictors

```{r}
ggplot(data.frame(pred = predictions, residuals = test_data$segment_time_s - predictions),
       aes(x = pred, y = residuals)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  ggtitle("Residuals vs Predictions")
```

### Polynomial Regression

#### Model Fit

```{r}
lm_poly <- lm(
  segment_time_s ~ segment_distance_m + I(segment_distance_m^2) + .,
  data = train_data
)

summary(lm_poly)
```

#### Evaluation

```{r}
predictions_poly <- predict(lm_poly, newdata = test_data)
```

##### Plots

```{r}
plot_predicted_vs_actual(test_data, predictions_poly)

rmse <- sqrt(mean((test_data$segment_time_s - predictions_poly)^2))
rmse
```

