---
title: "OznalProject"
output: html_document
---

# Predicting Travel Time of Bratislava MHD

## Selecting Data

For this project, a publicly available open dataset provided by the City of Bratislava was utilized, accessible at

<https://data.bratislava.sk/datasets/aba12fd2cbac4843bc7406151bc66106/about>

The dataset comprises raw data collected by the municipal public transport operator (MHD Bratislava) and is distributed across multiple text files. These files are interconnected through foreign key relationships, forming a structure analogous to a relational database. The version of the dataset employed in this study corresponds to the most recent update available at the time of analysis, released on April 17, 2026.

## Defining Hypothesis

The project is hypothesis-driven meaning a hypothesis will be evaluated at its end. The hypothesis is stated as follows:

*We hypothesize that scheduled travel time between two stops in GTFS-based public transport systems can be accurately predicted using a high-dimensional set of features (\>30 predictors), including spatial (e.g., Euclidean distance, geographic coordinates), temporal (e.g., arrival time, time-of-day transformations), route-specific, and stop-specific attributes. Furthermore, we hypothesize that incorporating engineered and encoded features (such as frequency-encoded categorical variables and interaction terms) will significantly improve predictive performance compared to using only basic spatial features.*

## Selecting Scenarios

Based on the hypothesis and data structure we decided to go with these two scenarios:

1.  (Scenario 1) Compare a regression or classification setup across three statistical or machine-learning methods and two feature-space partitioning approaches. Identify which features remain informative across both method families, discuss the implications for explainability, and outline the benefits each approach offers.
2.  (Scenario 3) Extend your work by comparing one algorithmic (forward, backward, mixed selection) and two embedded feature‑selection methods (lasso, ridge, elastic net) for the same regression or classification task. This task may require some research on your side, as not all necessary information will be taught through the course. Report how many features you ultimately retain, and which features lose statistical significance as you add or remove predictors.

## Analyzing Data

As introduced earlier, the dataset represents a General Transit Feed Specification (GTFS) system of public transportation operated by MHD Bratislava. The data are distributed across multiple text files, each capturing a different aspect of the transport system. The most relevant files used in this analysis include:

1.  *stops.txt* — contains attributes describing individual stops (e.g., Hlavná stanica, Pod stanicou), including their geographic coordinates.
2.  *trips.txt* — represents individual trips performed by vehicles, linking routes with sequences of stops.
3.  *stop_times.txt* — provides detailed timing information for each stop within a given trip.
4.  *routes.txt* — contains metadata about routes, such as route type and identifiers.
5.  Additional auxiliary files provided within the GTFS structure.

Several files and attributes are either sparsely populated or not utilized in this study due to missing or incomplete data. Examples include files such as *frequencies.txt* and *pathways.txt*, as well as attributes like *bike_allowed*, *stop_headsign*, and *shape_dist_traveled*. These were excluded to ensure data quality and consistency during modelling.

### Target - Regression or Classification?

The hypothesis states that we are predicting travel time which looks like a typical regression problem. But after analyzing the values of our target variable we got these results:

- 0 - 165 observations

- 15 - 180 observations

- 30 - 195 observations

- ...

- 540 - 2 observations

Although the target variable assumes discrete values (primarily multiples of 15 seconds) it represents an underlying continuous physical quantity, namely travel time. The variable is numeric and ordered, and differences between values are meaningful (e.g., 60 vs. 120 seconds). The observed discretization is a consequence of measurement precision or timetable rounding, rather than an inherent property of the phenomenon itself.

Given this, a regression-based approach is more appropriate than classification. Regression models preserve both the ordinal structure and magnitude of the target variable, enabling the modelling of continuous relationships between predictors and travel time. In contrast, treating the problem as classification would lead to a loss of information and reduced interpretability.

```         
```

## Experimental Setup

### Libraries

The analysis was conducted using the R programming language, with several libraries supporting data manipulation, visualization, and modelling. The *tidyverse* collection, including packages such as *dplyr*, *readr*, *ggplot2*, and *purrr*, was primarily used for data preprocessing, transformation, and visualization. The *patchwork* package was employed to combine multiple plots into unified visual outputs.

For statistical modelling and machine learning, multiple specialized libraries were utilized. Linear and diagnostic analyses were supported by the *car* package, while tree-based models were implemented using *rpart* (decision trees) and *ranger* (efficient random forest implementation). Regularized regression methods, including Lasso and Elastic Net, were implemented using the *glmnet* package

```{r}
library("tidyverse")
library("readr")
library("dplyr")

library("ggplot2")
library("patchwork")

library("purrr")
library("car")
library("rpart") # decision tree
library("ranger")
library("glmnet") # for lasso regression
library("rsample") # for cv
```

### Constants

The following constants are defined to ensure consistency and reproducibility across the project. The variable *TARGET* specifies the response variable to be predicted, while *SEED* is used to control randomness in operations such as data splitting and model training. The constant *CV_N_FOLDS* determines the number of folds used in cross-validation, enabling a robust evaluation of model performance.

```{r}
TARGET <- 'segment_time_s'
SEED <- 42
CV_N_FOLDS <- 5
SLICE_SAMPLE = 0.1
TRAIN_DF_PROPORTION = 0.8
```

## EDA

### Loading Data

Apart from loading dataset, we also perform **stratified_sampling** where we pick 10% of observations for each day_type. This ensures that each category is represented with respect to its proportion in data.

```{r}
timetable <- read_csv("../data/interim/dataset.csv")

table(timetable$day_type)
```

### Stratified Sampling

The dataset was reduced using a stratified sampling approach to ensure computational efficiency while preserving its original structure. Specifically, a 10% sample of unique trip_id values was selected within each category of the day_type variable, ensuring that the relative distribution of day types remained consistent.

```{r}
set.seed(SEED)

sampled_trips <- timetable %>%
  distinct(trip_id, day_type) %>%
  group_by(day_type) %>%
  slice_sample(prop = SLICE_SAMPLE) %>%
  ungroup() %>%
  pull(trip_id)

timetable <- timetable %>%
  filter(trip_id %in% sampled_trips)

table(timetable$day_type)
```

### Viewing Raw Data

Now we can take a lot at the dataset to see its structure and info about its columns.

```{r}
glimpse(timetable)
```

### Data Preparation

An initial exploratory inspection of the dataset revealed several data quality issues that needed to be addressed prior to modelling. These included the presence of missing or sparsely populated values, constant or near-constant features, uninformative attributes (such as URLs or descriptive text), and unsupported data types (e.g., raw strings or time formats not directly suitable for modelling).

To ensure compatibility with machine learning algorithms and improve model performance, these issues were systematically resolved during the data preparation phase. The objective of this step was to transform the raw dataset into a structured, consistent, and training-ready format suitable for subsequent analysis and modelling tasks.

#### Utility Functions

In this section we define a set of utility function that we are going to use in the later phases. Both of the functions do the same - they take a vector of columns names or row indices and drop it from the dataset. Apart from this, they also print some nice logs into console about the reduction.

```{r}
drop_cols_report <- function(data, cols_to_drop, label = NULL) {
  col_num_before <- ncol(data)
  
  data_reduced <- data %>%
    select(!any_of(cols_to_drop))
  
  col_num_after <- ncol(data_reduced)
  dropped_count <- col_num_before - col_num_after
  
  label_text <- if (!is.null(label)) paste0("[", label, "] ") else ""
  
  message(paste0(
    label_text,
    "Dropped cols: ", dropped_count,
    " | Current size: ", col_num_after
  ))
  
  return(data_reduced)
}

drop_rows_report <- function(data, rows_to_drop) {
  # 1. Store initial count
  row_num_before <- nrow(data)
  
  # 2. Perform the drop
  # Using -rows_to_drop handles the removal by index.
  # We check if rows_to_drop is empty first to avoid returning an empty table.
  if (length(rows_to_drop) > 0) {
    data_reduced <- data[-rows_to_drop, ]
  } else {
    data_reduced <- data
  }
  
  # 3. Calculate difference
  row_num_after <- nrow(data_reduced)
  dropped_count <- row_num_before - row_num_after
  
  # 4. Print the report
  message(paste0("Dropped rows: ", dropped_count, 
                 " - Current size: ", row_num_after))
  
  return(data_reduced)
}

find_duplicate_columns <- function(df) {
  is_dup <- duplicated(as.list(df))
  
  dup_summary <- tibble::tibble(
    column = names(df),
    is_duplicate = is_dup
  )
  
  return(dup_summary %>% filter(is_duplicate) %>% pull(column))
}

# =========================================================
# 1. One-hot encoding for low-cardinality predictors
#    Returns the original dataset with encoded columns added
#    and the original column removed.
# =========================================================

one_hot_encode <- function(data, col) {
  data[[col]] <- as.factor(data[[col]])
  
  mm <- model.matrix(
    as.formula(paste0("~ `", col, "`")),
    data = data
  ) %>%
    as.data.frame()
  
  # remove intercept column
  mm <- mm %>%
    dplyr::select(-`(Intercept)`)
  
  data %>%
    dplyr::bind_cols(mm) %>%
    dplyr::select(-all_of(col))
}

# =========================================================
# 2. Frequency encoding
#    Replaces each category by its relative frequency.
#    Good for medium/high-cardinality categoricals.
# =========================================================
frequency_encode <- function(data, col, new_col = NULL) {
  if (is.null(new_col)) {
    new_col <- paste0(col, "_freq")
  }
  
  freq_tbl <- data %>%
    count(.data[[col]], name = "n") %>%
    mutate(freq = n / sum(n)) %>%
    select(all_of(col), freq)
  
  data %>%
    left_join(freq_tbl, by = col) %>%
    rename(!!new_col := freq)
}

# =========================================================
# 3. Hex color -> RGB
#    Good for route_color if you want numeric representation
#    without category explosion.
# =========================================================
hex_to_rgb_df <- function(data, col = "route_color", prefix = "route") {
  hex_to_rgb <- function(hex) {
    hex <- gsub("#", "", as.character(hex))
    if (is.na(hex) || nchar(hex) != 6) return(c(NA, NA, NA))
    c(
      strtoi(substr(hex, 1, 2), 16L),
      strtoi(substr(hex, 3, 4), 16L),
      strtoi(substr(hex, 5, 6), 16L)
    )
  }
  
  rgb_vals <- t(sapply(data[[col]], hex_to_rgb)) %>%
    as.data.frame()
  
  colnames(rgb_vals) <- c(
    paste0(prefix, "_r"),
    paste0(prefix, "_g"),
    paste0(prefix, "_b")
  )
  
  bind_cols(data, rgb_vals)
}
```

#### Handling Missing Values

Throughout the following sections we are going to do the same:

1.  Create summaries about columns or rows which are somehow problematic
2.  Extract the column names or row indexes in the form of vectors

This structured format ensures that they can be effectively dropped by the reusable functions defined earlier.

In the code below, we create a summary of missing values for every columns using *is.na()* provided by base R. The summaries are then split into 2 vectors containing:

1. **Low-missing features** (0 < missing_pct < 50): These features are generally retained and handled using imputation techniques (e.g., mean, median, or model-based imputation), as sufficient information is available to preserve their predictive value.

2. **High-missing features** (missing_pct ≥ 50): These features are potentially problematic, as the large proportion of missing values may reduce their reliability. Such features are typically removed

```{r}
# 1. Missing values
# =========================

missing_summary <- timetable %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "missing_count") %>%
  mutate(missing_pct = missing_count / nrow(timetable) * 100) %>%
  filter(missing_pct > 0) %>%
  arrange(desc(missing_pct))

missing_summary
```

#### Constant & Near-Constant Features

These are features that carry no or only minimal signal and would only cause model to overfit they should be dropped too.

```{r}
# =========================
# 2. Constant & Near-Constant features
# =========================
variance_summary <- timetable %>%
  summarise(across(everything(), ~ n_distinct(.))) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "n_unique") %>%
  arrange(n_unique)
variance_summary
```

##### Constant Features

```{r}
# ===
# 2.1 Constant features (should be dropped immediately)
# ===

constant_features <- variance_summary %>%
  filter(n_unique == 1) %>% pull(column)

constant_features

```
##### Near-Constant Features
We identify near-constant features by calculating the ratio between the 2 largest frequencies identified in a column. If this ratio is above 20 we consider this column to be near-constant.

```{r}
# formula: ratio = freq_1 / freq_2
nzv_summary <- timetable %>%
  summarise(across(everything(), ~ {
    freq <- sort(table(.), decreasing = TRUE)
    if (length(freq) <= 1) return(NA)
    freq[1] / freq[2]
  })) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "freq_ratio") %>%
  arrange(desc(freq_ratio))

print(nzv_summary)


nzv_cols <- nzv_summary %>%
  filter(freq_ratio > 20) %>%
  pull(column)

print(nzv_cols)
```

#### Handling Duplicates

Now we also verify whether dataset doesn't contain duplicate rows (stored as vector of row indices) and duplicate columns (stored as vector of column names).

##### Duplicate Rows

```{r}
dup_row_indexes <- timetable %>%
  duplicated() %>%
  which()

print(paste("Duplicate row count:", length(dup_row_indexes)))

```

##### Duplicate Cols

The function `find_duplicate_columns` identifies duplicate columns by treating the dataset as a list of column vectors and applying the `duplicated()` function. This approach detects columns that have identical values to a previously encountered column.  It also includes edge cases such as columns containing only missing values (e.g., all `NA`), which are also considered duplicates if more than one such column exists.

```{r}
# More efficient check for duplicate columns
# R looks at the "top level" of the data frame. duplicated() now looks at each 
# vector (column) as a single item in a list and compares them.
dup_cols <- find_duplicate_columns(timetable)

paste("Duplicate cols")
dup_cols
```
##### Highly Correlated Features

Several features were identified as highly redundant due to their strong correlation with derived or engineered variables and were therefore removed from the dataset. In many cases, these original attributes served primarily as intermediate representations used to construct more informative features. For example, the *arrival_time* variable (stored in datetime format) was decomposed into components such as hour and minute, which better capture temporal patterns. Similarly, the *service_id* variable encodes predefined combinations of weekday indicators (*monday–sunday*), which were consolidated into a more interpretable categorical feature (*day_type*). The *shape_id* attribute was transformed into a continuous variable representing route geometry, namely *shape_distance*, computed as the cumulative Euclidean distance between stops. Additionally, identifiers such as *stop_id* were found to be redundant with other attributes (e.g., *stop_code*), and thus removed to reduce multicollinearity.

```{r}
highly_correlated_features = c('arrival_time', 'service_id', 'shape_id', 'stop_id',
                           'monday', 'tuesday', 'wednesday', 'thursday', 'friday',
                           'saturday', 'sunday', 'stop_sequence')
```


#### Feature Reduction

Problematic attributes, such as highly sparse features, constant or near-constant variables, duplicate columns, and duplicate observations, provide little or no useful information for model training. Therefore, these features and rows were removed. This resulted in a cleaner and more compact dataset suitable for subsequent preprocessing, feature engineering, and modelling.

```{r}
timetable_reduced <- drop_cols_report(timetable, high_missing, label = "high missing")
timetable_reduced <- drop_cols_report(timetable_reduced, constant_features, label = "constant features")
timetable_reduced <- drop_cols_report(timetable_reduced, nzv_cols, label = "near-constan features")
timetable_reduced <- drop_rows_report(timetable_reduced, dup_row_indexes)
timetable_reduced <- drop_cols_report(timetable_reduced, dup_cols, label = "duplicate cols")
timetable_reduced <- drop_cols_report(timetable_reduced, highly_correlated_features, label = "highly_correlated")

glimpse(timetable_reduced)
```

#### Handling low-missing features

Let's first analyze the low-missing features:

```{r}
missing_summary %>% filter(missing_pct < 50 & missing_pct > 0)
```

##### time_since_last_stop

This one is calculated as:

*time_since_last_stop = arrival_seconds - prev(departure_seconds)*

where *prev()* represents the previous stop within a trip (sorted by *stop_sequence*). 

```{r}
summary(timetable_reduced$time_since_last_stop)
```
The NaN values logically represent the first stop within a trip which doesn't have a previous stop. Thus, we apply imputation with 0:
```{r}

timetable_imputed <- timetable_reduced %>%
  mutate(
    time_since_last_stop = replace_na(time_since_last_stop, 0)
  )
```

##### wheelchair_accessible

Since wheelchair_accessible mostly contains value 1 (which means accessible) and a very small portion of missing values it cannot serve as a reliable predictor and should be dropped too.

```{r}
table(timetable_imputed$wheelchair_accessible)
timetable_imputed <- drop_cols_report(timetable_imputed, c('wheelchair_accessible'))
```

#### Feature Encoding

Now, the remaining preprocessing step focuses on encoding categorical and non-ordinal features. Three main techniques were applied:

1.  **One-hot encoding**: Used for categorical features with relatively low cardinality (typically fewer than 15 unique values) to preserve interpretability while avoiding excessive dimensionality.
2.  **Frequency encoding**: Applied to high-cardinality categorical features, where one-hot encoding would lead to a large number of sparse columns. This approach replaces each category with its relative frequency, retaining distributional information in a compact form.
3.  **Feature transformation and decomposition**: Certain features were transformed into more informative numerical representations. For example, hexadecimal color codes were decomposed into RGB components.

In contrast to one-hot encoding, frequency encoding is applied after the train/test split to prevent data leakage. Specifically, frequency mappings are learned exclusively from the training data and then applied to both training and testing sets. This ensures that information from the test set does not influence the encoding process, thereby preserving the integrity of model evaluation and avoiding overly optimistic performance estimates.

```{r}
glimpse(timetable_imputed)

timetable_encoded <- timetable_imputed %>%
  one_hot_encode("platform_code") %>%
  one_hot_encode("day_type") %>%
  one_hot_encode("zone_id") %>%
  one_hot_encode("direction_id") %>%
  one_hot_encode("pickup_type") %>%
  one_hot_encode("route_type") %>%
  hex_to_rgb_df("route_color", prefix = "route") %>%
  select(
    -route_color,
  )

glimpse(timetable_encoded)
```

#### Saving Dataset

```{r}
# =========================
# 10. Save dataset
# =========================
saveRDS(timetable_encoded, "./datasets/timetable_encoded.rds")
```

### Training Preparation

Now that issues such as missing values, constant or near-constant features, and appropriate encoding have been addressed, the resulting dataset represents a set of candidate predictors suitable for model training.

In this second phase of exploratory data analysis, the focus shifts from general data quality to model-specific requirements. In particular, we examine:

1.  **Multicollinearity** among predictors, which can negatively affect linear models
2.  **Extreme outliers** in feature distributions, which may influence model stability
3.  **The nature of relationships** between predictors and the target variable, including potential non-linear patterns

To support this analysis, various visualization techniques such as heatmaps, boxplots, and scatterplots are employed using libraries like ggplot2. These tools help to better understand how the data aligns with the assumptions and behavior of different regression models.

Based on these insights, appropriate transformations and feature selection strategies are proposed to improve model performance and robustness.

#### Loading Data

```{r}
timetable_encoded <- readRDS("./datasets/timetable_encoded.rds")
glimpse(timetable_encoded)
```

#### Utility Functions

This section contains utility functions that are going to be repeteadly used throughout this phase.

```{r}

select_correlated_features <- function(
    data,
    target_col,
    min_abs_cor = 0.1,
    top_n = NULL,
    always_keep = NULL
) {
  # keep only numeric columns for correlation
  numeric_data <- data %>%
    dplyr::select(where(is.numeric))
  
  # check target exists
  if (!(target_col %in% names(numeric_data))) {
    stop(paste("Target column", target_col, "must exist and be numeric."))
  }
  
  # compute correlations with target
  cor_vec <- cor(numeric_data, use = "complete.obs")[, target_col]
  
  cor_df <- data.frame(
    feature = names(cor_vec),
    correlation = as.numeric(cor_vec)
  ) %>%
    dplyr::filter(feature != target_col) %>%
    dplyr::mutate(abs_correlation = abs(correlation)) %>%
    dplyr::arrange(desc(abs_correlation))
  
  # threshold filter
  cor_df <- cor_df %>%
    dplyr::filter(abs_correlation >= min_abs_cor)
  
  # optional top_n restriction
  if (!is.null(top_n)) {
    cor_df <- cor_df %>%
      dplyr::slice_head(n = top_n)
  }
  
  # final feature list
  selected_features <- unique(c(
    target_col,
    cor_df$feature,
    always_keep
  ))
  
  # keep only existing columns (safe)
  selected_features <- intersect(selected_features, names(data))
  
  return(data %>%
           dplyr::select(all_of(selected_features)))
}

plot_correlation_heatmap <- function(df, threshold = 0.3, title = "Correlation Heatmap") {
  
  # 1. Calculate correlation matrix (only for numeric columns)
  # It's safer to filter for numeric variables automatically
  cor_matrix <- df %>%
    select(where(is.numeric)) %>%
    cor(use = "complete.obs")
  
  # 2. Convert to long format
  cor_long <- as.data.frame(as.table(cor_matrix))
  colnames(cor_long) <- c("Var1", "Var2", "Correlation")
  
  # 3. Create the plot
  ggplot(cor_long, aes(x = Var1, y = Var2, fill = Correlation)) +
    geom_tile() +
    # Dynamic label filtering based on the 'threshold' argument
    geom_text(
      aes(label = ifelse(abs(Correlation) > threshold, round(Correlation, 2), "")),
      size = 3,
      color = "black"
    ) +
    scale_fill_gradient2(
      low = "#377eb8",  # Nice blue
      mid = "white",
      high = "#e41a1c", # Nice red
      midpoint = 0,
      limit = c(-1, 1)   # Ensures the scale is always centered
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.title = element_blank(),
      panel.grid = element_blank()
    ) +
    ggtitle(title)
}

plot_hist_dist <- function(df, var_name, bins = 30, show_density = FALSE) {
  p <- ggplot(df, aes(x = !!sym(var_name)))
  
  if (show_density) {
    # If density is requested, we must map the y-axis to density
    p <- p + geom_histogram(
      aes(y = after_stat(density)),
      fill = "steelblue", 
      color = "white", 
      bins = bins, 
      alpha = 0.6
    ) +
    geom_density(color = "darkblue", linewidth = 1)
  } else {
    # Standard count-based histogram
    p <- p + geom_histogram(
      fill = "steelblue", 
      color = "white", 
      bins = bins, 
      alpha = 0.8
    )
  }
  
  p + theme_minimal() +
    labs(
      title = paste("Distribution of", var_name),
      x = var_name,
      y = if(show_density) "Density" else "Frequency"
    )
}

# plot_hist_dist <- function(df, var_name, bins = 30) {
#   ggplot(df, aes(x = !!sym(var_name))) +
#     geom_histogram(
#       fill = "steelblue", 
#       color = "white", 
#       bins = bins, 
#       alpha = 0.8
#     ) +
#     theme_minimal() +
#     labs(
#       title = paste("Distribution of", var_name),
#       x = var_name,
#       y = "Frequency"
#     )
# }

plot_box_dist <- function(df, var_name) {
  ggplot(df, aes(x = !!sym(var_name))) +
    # Adding a dummy y-axis value or a factor(0) helps boxplot orientation
    geom_boxplot(fill = "lightblue", outlier.color = "red", outlier.alpha = 0.5) +
    coord_flip() +
    theme_minimal() +
    labs(
      title = paste(var_name),
      x = NULL # Removing x label as the title covers it
    )
}

plot_predictor_boxplots <- function(
    df,
    target_col = "segment_time_s",
    ncol = 3,
    show_message = TRUE,
    title = "Outlier Detection Across Predictors",
    subtitle = "Boxplots identifying potential outliers in red"
) {
  # Identify binary columns
  is_binary <- purrr::map_lgl(df, ~ length(unique(stats::na.omit(.x))) <= 2)
  binary_cols <- names(is_binary[is_binary])
  
  # Select predictors to plot
  predictors_to_plot <- setdiff(names(df), c(target_col, binary_cols))
  
  # Optional message
  if (show_message) {
    message("Dropping binary features: ", paste(binary_cols, collapse = ", "))
  }
  
  # Create individual plots
  plot_list <- purrr::map(predictors_to_plot, ~ plot_box_dist(df, .x))
  
  # Combine plots
  combined_plot <- patchwork::wrap_plots(plot_list, ncol = ncol) +
    patchwork::plot_annotation(
      title = title,
      subtitle = subtitle,
      theme = ggplot2::theme(
        plot.title = ggplot2::element_text(size = 18, face = "bold")
      )
    )
  
  return(combined_plot)
}

# --- Engineering --- #


add_engineered_features <- function(df){
  df_engineered <- df %>%
    mutate(
      sin_time = sin(2 * pi * seconds_since_midnight / 86400),
      cos_time = cos(2 * pi * seconds_since_midnight / 86400),
      distance_sin_interaction = segment_distance_m * sin_time,
      distance_cos_interaction = segment_distance_m * cos_time,
    )
  
  return(df_engineered)
}

```

#### Target Distribution

Now we can also analyze the distribution of our target variable

```{r}

plot_hist_dist(timetable_encoded, TARGET, bins = 30, show_density = FALSE)
```

Most observations are concentrated around **60 seconds**, with a gradual tail toward higher values. This indicates that **typical segment travel times are short**, while longer durations occur less frequently but extend the upper range.

#### Data Engineering

In order to better capture temporal patterns present in the data, the original time variable was transformed into a cyclical representation. Specifically, the variable seconds_since_midnight was mapped onto the unit circle using sine and cosine transformations. This approach preserves the periodic nature of time, ensuring that values close in real-world time (e.g., late night and early morning) are also close in the transformed feature space.

In addition to temporal encoding, interaction features were introduced to model the combined effect of distance and time of day. The features distance_sin_interaction and distance_cos_interaction were created by multiplying the segment distance with the corresponding sine and cosine components. These interaction terms allow the model to capture time-dependent variations in travel behavior, such as differing traffic conditions across the day.

```{r}
timetable_engineered <- add_engineered_features(timetable_encoded)

# timetable_engineered <- timetable_encoded %>%
#   mutate(
#     sin_time = sin(2 * pi * seconds_since_midnight / 86400),
#     cos_time = cos(2 * pi * seconds_since_midnight / 86400),
#     distance_sin_interaction = segment_distance_m * sin_time,
#     distance_cos_interaction = segment_distance_m * cos_time,
#   )

glimpse(timetable_engineered)
```

##### Visualizations

Now let's visualize the engineered features' distributions by plotting histograms:

```{r}
timetable_engineered %>%
  select(sin_time, cos_time, distance_sin_interaction, distance_cos_interaction) %>%
  pivot_longer(cols = everything(), names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  facet_wrap(~ feature, scales = "free") +
  theme_minimal()
```

The distributions of the cyclical features *sin_time* and *cos_time* exhibit smooth, symmetric patterns consistent with their trigonometric formulation. Both variables show higher densities near their extreme values (−1 and 1), with a gradual decline toward intermediate values, followed by a mirrored increase.

The interaction features, *distance_sin_interaction* and *distance_cos_interaction*, are centered around zero and exhibit approximately bell-shaped distributions, suggesting a near-normal structure. However, the presence of heavy tails reflects the influence of large distance values, resulting in more extreme interaction effects.

To verify the distributions of *sin_time* and *cos_time* even further we can do **circular structure check**.

```{r}
ggplot(timetable_engineered, aes(x = sin_time, y = cos_time)) +
  geom_point(alpha = 0.1) +
  coord_fixed() +
  theme_minimal()
```

The joint visualization of these features confirms their correct transformation, forming a well-defined circular structure, which indicates that the periodic nature of time has been successfully preserved.

#### Checking multicollinearity & target collinearity

One of the key assumptions of linear and polynomial regression models is the absence of strong multicollinearity among predictor variables, as highly correlated features can distort coefficient estimates and reduce model interpretability. To assess this, a correlation matrix was computed and visualized using a heatmap, allowing for a clear inspection of pairwise relationships between variables.

The results indicate several strong correlations among features, particularly between *cos_time* and *distance_cos_interaction* , as well as between *sin_time* and *distance_sin_interaction*. These relationships are expected, since the interaction features are directly derived from their respective base variables. Additionally, strong negative correlations are observed between temporal sine/cosine components and variables such as *seconds_since_midnight* and *arrival_seconds*, reflecting their cyclical encoding.

```{r}
correlated_features <- c('cos_time', 'distance_cos_interaction', 'sin_time', 'distance_sin_interaction',
                         'arrival_seconds', 'seconds_since_midnight')

plot_correlation_heatmap(
  timetable_engineered %>% 
    select(all_of(c(correlated_features, "segment_time_s"))) %>%
    select(-all_of(c(TARGET)))
)
```

#### Outliers

Extreme outliers are also problematic for regression models. Let's visualize them using boxplots.

##### Boxplots

```{r}

skewed_features <- c('segment_distance_m', 'distance_sin_interaction',
                          'distance_cos_interaction')

combined_plot <- plot_predictor_boxplots(timetable_engineered %>% select(all_of(skewed_features)))
print(combined_plot)
```

##### Transformations

Based on the analysis of the predictors with most outliers and their distributions, we recommend following before using them in training:

1.  *segment_distance_m* -\> very long tail, strong right skew, **log_transform** recommended
2.  *distance_interactions* -\> also heavy tails, caused by distance metric. Log cannot be applied for negative values, so transforming distance and the recomputing is recommended

Log transformation was applied to reduce skewness and stabilize variance, improving the suitability of predictors for linear modelling. Alternative transformations such as Box-Cox or Yeo-Johnson could also be considered; however, log transformation provides a simple and effective solution for positively skewed variables.

### Saving Data

```{r}
saveRDS(timetable_engineered, './datasets/timetable_engineered.rds')
```

## Model Training

### Utility Functions

```{r}
# ---- ALTERNATIVE (LESS VISUALLY APPEALING) ----
# plot_homoscedasticity_plot <- function(fitted, residuals){
#   data.frame(
#     fitted = fitted,
#     residuals = residuals
#   ) %>%
#     ggplot(aes(x = fitted, y = residuals)) +
#     geom_point(alpha = 0.2) +
#     geom_hline(yintercept = 0, color = "red") +
#     theme_minimal()
# }
plot_homoscedasticity_plot <- function(fitted, residuals){
  data.frame(
    fitted = fitted,
    residuals = residuals
  ) %>%
    ggplot(aes(x = fitted, y = residuals)) +
    
    # points
    geom_point(alpha = 0.1, size = 0.6, color = "steelblue") +
    
    # horizontal zero line
    geom_hline(yintercept = 0, color = "red", linewidth = 0.8) +
    
    # smooth trend (VERY IMPORTANT)
    geom_smooth(
      method = "gam",
      formula = y ~ s(x, k = 5),
      color = "orange",
      se = FALSE,
      linewidth = 1
    ) +
    
    # optional: highlight spread
    stat_density_2d(alpha = 0.2, color = "grey50") +
    
    theme_minimal() +
    labs(
      title = "Residuals vs Fitted Values",
      subtitle = "Checking homoscedasticity",
      x = "Fitted values",
      y = "Residuals"
    ) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
}

plot_scale_location <- function(fitted, residuals){
  data.frame(
    fitted = fitted,
    sqrt_abs_resid = sqrt(abs(residuals))
  ) %>%
    ggplot(aes(x = fitted, y = sqrt_abs_resid)) +
    geom_point(alpha = 0.1) +
    geom_smooth(method = "gam", se = FALSE, color = "blue") +
    theme_minimal() +
    labs(
      title = "Scale-Location Plot",
      y = "√|Residuals|"
    )
}

# --- ALTERNATIVE TO BINNED RELATIONSHIP PLOT ---
# plot_scatterplot_line <- function(data, x, y = "segment_time_s") {
#   ggplot(data, aes(x = .data[[x]], y = .data[[y]])) +
#     geom_point(alpha = 0.08, size = 0.6) +
#     geom_smooth(method = "lm", color = "red", se = FALSE, linewidth = 0.8) +
#     geom_smooth(
#       method = "gam",
#       formula = y ~ s(x, k = 5),
#       color = "blue",
#       se = FALSE,
#       linewidth = 0.9
#     ) +
#     theme_minimal() +
#     labs(
#       title = x,
#       x = x,
#       y = y
#     ) +
#     theme(
#       plot.title = element_text(size = 10, face = "bold"),
#       axis.text = element_text(size = 8),
#       axis.title = element_text(size = 8)
#     )
# }
# scatterplot_list <- purrr::map(
#   features_to_check,
#   ~ plot_scatterplot_line(sample_data, .x)
# )
# 
# combined_plot <- patchwork::wrap_plots(scatterplot_list, ncol = 2)
# 
# combined_plot

plot_binned_relationship <- function(data, x, y = "segment_time_s", bins = 30, 
                                     label_size = 12, text_size = 10) {
  data %>%
    mutate(bin = ntile(.data[[x]], bins)) %>%
    group_by(bin) %>%
    summarise(
      x_mean = mean(.data[[x]], na.rm = TRUE),
      y_mean = mean(.data[[y]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    ggplot(aes(x = x_mean, y = y_mean)) +
    geom_point(size = 1.2, alpha = 0.4) +
    geom_smooth(aes(color = "Linear"), method = "lm", se = FALSE) +
    geom_smooth(
      aes(color = "Non-linear"),
      method = "gam",
      formula = y ~ s(x, k = 5),
      se = FALSE
    ) +
    scale_color_manual(
      name = "Legend", # This name must be identical across all plots
      values = c("Linear" = "red", "Non-linear" = "blue")
    ) +
    theme_minimal() +
    theme(
      axis.title = element_text(size = label_size),
      axis.text = element_text(size = text_size)
    ) +
    labs(x = x, y = paste("Mean", y))
}

# ---- Evaluation ---- #

evaluate_regression <- function(actual, predicted, round_to = NULL, model_name = "model") {
  if (!is.null(round_to)) {
    predicted <- round(predicted / round_to) * round_to
  }
  
  rmse <- sqrt(mean((actual - predicted)^2, na.rm = TRUE))
  mae  <- mean(abs(actual - predicted), na.rm = TRUE)
  
  ss_total <- sum((actual - mean(actual, na.rm = TRUE))^2, na.rm = TRUE)
  ss_res   <- sum((actual - predicted)^2, na.rm = TRUE)
  r2 <- 1 - (ss_res / ss_total)
  
  tibble::tibble(
    model = model_name,
    RMSE = rmse,
    MAE = mae,
    R2 = r2
  )
}

# explicitly used in cv
eval_metrics <- function(actual, predicted) {
  tibble(
    RMSE = sqrt(mean((actual - predicted)^2)),
    MAE = mean(abs(actual - predicted)),
    R2 = cor(actual, predicted)^2
  )
}

# used for LR, PR, DT, RF
eval_train_test <- function(model, model_label, train_data, test_data, predict_fun) {
  pred_train <- predict_fun(model, train_data)
  pred_test  <- predict_fun(model, test_data)
  
  dplyr::bind_rows(
    evaluate_regression(
      actual = train_data$segment_time_s,
      predicted = pred_train,
      model_name = paste0(model_label, " train")
    ),
    evaluate_regression(
      actual = test_data$segment_time_s,
      predicted = pred_test,
      model_name = paste0(model_label, " test")
    )
  )
}

# used for LASSO/ElASTIC
eval_train_test_glmnet <- function(model, model_label, x_train, y_train, x_test, y_test) {
  pred_train <- as.vector(predict(model, newx = x_train))
  pred_test  <- as.vector(predict(model, newx = x_test))
  
  dplyr::bind_rows(
    evaluate_regression(
      actual = y_train,
      predicted = pred_train,
      model_name = paste0(model_label, " train")
    ),
    evaluate_regression(
      actual = y_test,
      predicted = pred_test,
      model_name = paste0(model_label, " test")
    )
  )
}

plot_predicted_vs_actual <- function(
  data,
  predictions,
  target = "segment_time_s",
  title = "Predicted vs Actual Values",
  add_smooth = TRUE
) {
  plot_data <- data.frame(
    actual = data[[target]],
    predicted = predictions
  )
  
  # Create perfect line data
  line_data <- data.frame(
    x = range(plot_data$actual)
  )
  line_data$y <- line_data$x
  
  p <- ggplot(plot_data, aes(x = actual, y = predicted)) +
    
    geom_point(alpha = 0.25, size = 1, color = "steelblue") +
    
    # Perfect prediction line (now mapped → legend works)
    geom_line(
      data = line_data,
      aes(x = x, y = y, color = "Perfect prediction"),
      linewidth = 1,
      linetype = "dashed"
    )
  
  if (add_smooth) {
    p <- p + geom_smooth(
      aes(color = "Model trend"),
      method = "gam",
      formula = y ~ s(x, k = 5),
      se = FALSE,
      linewidth = 1
    )
  }
  
  p +
    scale_color_manual(
      name = "Lines",
      values = c(
        "Perfect prediction" = "red",
        "Model trend" = "darkorange"
      )
    ) +
    coord_equal() +
    theme_minimal() +
    labs(
      title = title,
      subtitle = "Comparison of predictions vs actual values",
      x = "Actual value",
      y = "Predicted value"
    ) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "bottom"
    )
}

plot_qq <- function(residuals, title = "QQ Plot") {
  data.frame(residuals = residuals) %>%
    ggplot(aes(sample = residuals)) +
    stat_qq(alpha = 0.3) +
    stat_qq_line(color = "red") +
    theme_minimal() +
    labs(
      title = title,
      subtitle = "Check of residual normality"
    )
}

# ---- Feature Selection ---- #

select_model_features <- function(data, features = NULL, target = "segment_time_s") {
  
  # If NULL or empty → return full dataset
  if (is.null(features) || length(features) == 0) {
    return(data)
  }
  
  keep_cols <- unique(c(target, features))
  
  missing_cols <- setdiff(keep_cols, names(data))
  if (length(missing_cols) > 0) {
    warning("Missing columns: ", paste(missing_cols, collapse = ", "))
  }
  
  data %>%
    dplyr::select(dplyr::any_of(keep_cols))
}

# ---- ENCODING ----- #

fit_frequency_encoding <- function(train_df, col) {
  freq_col <- paste0(col, "_freq")
  
  freq_table <- train_df %>%
    dplyr::count(.data[[col]])
  
  names(freq_table)[names(freq_table) == "n"] <- freq_col
  
  freq_table[[freq_col]] <- freq_table[[freq_col]] / nrow(train_df)
  
  freq_table
}

apply_frequency_encoding <- function(df, freq_table, col) {
  freq_col <- paste0(col, "_freq")
  
  df %>%
    dplyr::left_join(freq_table, by = col) %>%
    dplyr::mutate(
      !!freq_col := ifelse(is.na(.data[[freq_col]]), 0, .data[[freq_col]])
    ) %>%
    dplyr::select(-dplyr::all_of(col))
}


# ---- Lasso & Elastic Helpers ---- #

prepare_glmnet_data <- function(train_data, test_data, target) {
  feature_names <- setdiff(names(train_data), target)
  
  x_train <- as.matrix(train_data[, feature_names])
  y_train <- train_data[[target]]
  
  x_test <- as.matrix(test_data[, feature_names])
  y_test <- test_data[[target]]
  
  list(
    x_train = x_train,
    y_train = y_train,
    x_test = x_test,
    y_test = y_test,
    feature_names = feature_names
  )
}

# ---- VIF Helpers (LR, PR) ----

format_vif <- function(vif_values) {
  data.frame(
    feature = names(vif_values),
    vif = as.numeric(vif_values)
  ) %>%
    dplyr::arrange(desc(vif))
}
```

### Loading Data

The predictor `time_since_last_stop` was removed prior to model training, as it may introduce data leakage by incorporating temporal information that is not available at prediction time. However, this feature was retained during the exploratory data analysis phase, where it was frequently used for understanding relationships within the data. To avoid disrupting the consistency and completeness of the EDA process, the feature was only excluded at the modeling stage, ensuring both valid analysis and reliable model evaluation.

```{r}
timetable <- readRDS("./datasets/timetable_engineered.rds")


features_to_drop <- c('time_since_last_stop')
timetable <- drop_cols_report(timetable, features_to_drop)

```

### Transformations

During exploratory data analysis, several predictors were identified as exhibiting highly skewed distributions, namely *segment_distance_m*, *distance_sin_interaction*, and *distance_cos_interaction*. To address this, a logarithmic transformation was applied selectively.

The transformation was performed only on *segment_distance_m*, as the remaining two features are derived from it through interaction with cyclical time components (i.e., distance multiplied by sine and cosine transformations of time). Additionally, logarithmic transformation is not directly applicable to these derived features due to the presence of negative values. Instead, after transforming the base distance variable, the interaction features were recomputed, ensuring consistency while preserving their interpretability.

```{r}


problematic_features <- c('segment_distance_m', 'distance_sin_interaction',
                          'distance_cos_interaction')

timetable_transformed <- timetable %>%
  mutate(
    segment_distance_m = log1p(segment_distance_m)
  )

timetable_transformed <- add_engineered_features(timetable_transformed)

plot_predictor_boxplots(timetable_transformed %>% select(problematic_features), target_col = 'segment_time_s')
```

From the box plots above it is clear that the skewness has been successfully eliminated for all the problematic features.


### Saving Training-Ready Data

```{r}
saveRDS(timetable_transformed, './datasets/training_ready_data.rds')
```

### Train/Test Split, Frequency Encoding & Folds
The function `create_train_test_split` is designed to prepare training and testing datasets in a consistent and reproducible manner and is applied uniformly across all models used in this study. The splitting procedure is performed at the *trip level* rather than on individual rows, ensuring that all observations belonging to a given trip are assigned exclusively to either the training or testing set. This approach prevents data leakage that could arise from having parts of the same trip present in both subsets.

The function begins by randomly sampling a proportion of unique *trip_id* values (80% for training, 20% for testing), effectively performing a grouped sampling strategy that preserves the structural dependencies within trips. A sanity check is then conducted to verify that no trip appears in both subsets.

Following the split, categorical variables are encoded using frequency encoding. Importantly, the encoding mappings are learned solely from the training data and subsequently applied to the test data, ensuring that no information from the test set influences the transformation process. This design choice further prevents data leakage and maintains the validity of model evaluation.

Finally, duplicate columns are identified and removed based on the training data, and the same columns are consistently dropped from the test data to ensure alignment between both datasets. The function returns the processed training and testing sets, which are then used as inputs for all subsequent modelling approaches.

```{r}

create_train_test_split <- function(data, seed = 42, train_size = TRAIN_DF_PROPORTION) {
  set.seed(seed)
  
  # 1. Vyber 80% trip_id (nie riadkov)
  unique_trips <- unique(data$trip_id)
  train_trips <- sample(unique_trips, size = train_size * length(unique_trips))
  
  # 2. Filter podla trip_id - cely trip ide bud do trainu, alebo do testu
  train_data <- data %>%
    filter(trip_id %in% train_trips)
  
  test_data <- data %>%
    filter(!trip_id %in% train_trips)
  
  # 3. Sanity check - ziadny trip nesmie byt v oboch
  shared_trips <- intersect(unique(train_data$trip_id), unique(test_data$trip_id))
  stopifnot(length(shared_trips) == 0)
  
  cat("Train trips:", length(unique(train_data$trip_id)), 
      "| Test trips:", length(unique(test_data$trip_id)), "\n")
  cat("Train rows:", nrow(train_data), 
      "| Test rows:", nrow(test_data), "\n")
  
  # 4. Az teraz dropnut trip_id - v modeli ho nechces
  train_data <- train_data %>% select(-trip_id)
  test_data  <- test_data  %>% select(-trip_id)
  
  freq_cols <- c(
    "route_short_name",
    "route_id",
    "trip_headsign",
    "stop_name",
    "stop_code"
  )
  
  # glimpse(train_data)
  
  for (col in freq_cols) {
    freq_table <- fit_frequency_encoding(train_data, col)
    
    train_data <- apply_frequency_encoding(train_data, freq_table, col)
    test_data  <- apply_frequency_encoding(test_data, freq_table, col)
  }
  
  dup_cols <- find_duplicate_columns(train_data)
  train_data <- drop_cols_report(train_data, dup_cols, label = paste0("train - duplicate col "))
  
  test_data <- drop_cols_report(test_data, dup_cols, label = paste0("test - duplicate col"))
                                 
  
  
  list(train = train_data, test = test_data, dup_cols = dup_cols)
}

create_folds <- function(train_data, v = CV_N_FOLD, seed = SEED){
  folds <- vfold_cv(train_data, v = v)
}
```

### Backward Elimination

Backward elimination was selected as the primary feature selection method due to its ability to evaluate predictors in the context of the full model. The method starts with all features and iteratively removes variables based on the Akaike Information Criterion (AIC), balancing model fit and complexity, and thus helping to prevent overfitting while maintaining predictive performance.

However, backward elimination does not explicitly address multicollinearity, which can lead to unstable coefficient estimates in regression models. Therefore, an additional feature selection step based on the Variance Inflation Factor (VIF) is applied for linear and polynomial regression models. This allows the removal of highly collinear predictors, improving model interpretability and stability.

### Linear Regression

Linear regression was chosen as the initial modeling approach due to its high interpretability and simplicity. As a baseline model, it provides a clear understanding of the relationships between predictors and the target variable through easily interpretable coefficients. This makes it a useful reference point for evaluating more complex models, allowing us to assess whether additional complexity leads to meaningful improvements in predictive performance.

#### Train/Test Split & Folds

```{r}
split <- create_train_test_split(timetable_transformed, seed = SEED)

train_data <- split$train
test_data  <- split$test


folds <- create_folds(train_data, v = CV_N_FOLDS, seed = SEED)
```

#### Checking Relationships

We are going to check the relationships between some of the most useful predictors and target. For the sake of performance, we are going to limit the total size of the dataset. For the sake of clarity, we are going to use binned scatterplots that:

1.  Split into x and y axis to a limited number of bins
2.  For each bin an average is computed
3.  For each pair a point is added

This approach notably limits the total number of points in the plot and improves overall clarity. Apart from points, the two lines are dragged through the space:

1.  linear (straight) line which indicates linear relationship
2.  non-linear (curved) line which indicates polynomial relationship

If both lines significantly overlap, it indicates linear relationship.

```{r}
sample_data <- train_data %>%
  slice_sample(n = 5000)

train_col_names <- colnames(sample_data)

train_col_names <- train_col_names[!train_col_names %in% c('segment_time_s')]

length(train_col_names)

colnames(sample_data)

features_to_check <- c(
  "segment_distance_m",
  "distance_cos_interaction",
  "distance_sin_interaction",
  "arrival_seconds",
  "arrival_time_hour",
  "arrival_time_minute",
  "arrival_time_second",
  "seconds_since_midnight",
  "stop_lat",
  "stop_lon"
)

# Note: Ensure TARGET is a string, e.g., TARGET <- "my_column_name"
binned_plot_list <- purrr::map(
  features_to_check,
  ~ plot_binned_relationship(sample_data, .x, y = TARGET, bins = 30, label_size = 8, text_size = 8)
)

# guides = "collect": Patchwork looks at all the plots. If it sees that they all have a legend named "Legend" with the same colors, it removes them from the individual plots and creates one single master legend.
patchwork::wrap_plots(binned_plot_list, ncol = 2, guides = "collect") & 
  theme(legend.position = "bottom")
```

Several predictors (especially interaction terms and geometric metrics) demonstrate clear **non-linear relationships with the target**, indicating that linear models alone may be insufficient and that incorporating non-linear transformations or more flexible models is justified.

#### Feature Selection

The features listed in `high_vif_features` were identified through an iterative process. First, a regression model was fitted using backward elimination based on the AIC criterion. Subsequently, the Variance Inflation Factor (VIF) was computed to assess multicollinearity among the remaining predictors. Features exhibiting high VIF values were then removed, and the model was refitted. This process was repeated until an acceptable level of multicollinearity was achieved.

You can skip this if you wish to keep the correlated features and use only pure backward elimination approach.

```{r}

high_vif_features <- c('sin_time', 'cos_time', 'route_type11',
'day_typesunday', 'arrival_time_hour', 'arrival_seconds')

train_data <- drop_cols_report(train_data, high_vif_features)
test_data <- drop_cols_report(test_data, high_vif_features)

```

#### Model Fit

```{r}

lm_model <- lm(segment_time_s ~ ., data = train_data)
initial_vars <- attr(terms(lm_model), "term.labels")

lm_step <- step(
  lm_model,
  direction = "backward",
  trace = FALSE
)
final_vars   <- attr(terms(lm_step), "term.labels")


final_lm_formula <- formula(lm_step)
# final_lm_formula

summary(lm_step)
```
#### Features Dropped

Here we list features that were dropped by Backward Elimination Approach.

```{r}
dropped_vars <- setdiff(initial_vars, final_vars)
kept_vars    <- final_vars

dropped_vars
```

#### VIF

The Variance Inflation Factor (VIF) analysis indicates that **multicollinearity is not a significant concern** in the final linear model. The majority of predictors exhibit very low VIF values, typically close to 1, suggesting minimal linear dependence among features.

```{r}

vif_values <- vif(lm_step)
vif_df <- format_vif(vif_values)
vif_df

```


#### Evaluation

```{r}
predict_lm <- function(model, new_data) {
  predict(model, newdata = new_data)
}

results_lm <- eval_train_test(
  lm_step,
  "Linear Regression",
  train_data,
  test_data,
  predict_fun = predict_lm
)

results_lm
```

##### Predicted vs. Actual

The plot shows a clear positive relationship between predicted and actual values, indicating that the linear regression model captures the overall trend in the data. However, the predictions are consistently compressed toward the mean, as evidenced by the model trend line being flatter than the perfect prediction line. This suggests that the model tends to underestimate higher values and overestimate lower values, indicating limited ability to capture extreme observations. Additionally, the noticeable spread around the trend line reflects residual variance, implying that while the model performs reasonably well, there is still unexplained variability in the data.

```{r}
plot_predicted_vs_actual(
  data = test_data,
  predictions = predict_lm(lm_step, test_data),
  target = "segment_time_s",
  title = "Linear Regression: Predicted vs Actual"
)
```

##### Homoscedasticity

The scale-location plots reveal that the linear regression model exhibits clear heteroscedasticity, particularly for higher fitted values, where the variance of residuals increases substantially. This indicates that the model fails to adequately capture the underlying structure of the data.

```{r}
plot_scale_location(lm_step$fitted.values, lm_step$residuals)
```

##### Normality of Residuals

The QQ plot compares the distribution of residuals from the linear regression model to a theoretical normal distribution. While the points closely follow the reference line in the central region, indicating approximate normality for the majority of residuals, there are clear deviations in the upper tail. Specifically, the points curve sharply upward, suggesting the presence of heavy right tails and outliers.

This indicates that the residuals are not perfectly normally distributed, with the model underestimating extreme values. Such deviations imply a violation of the normality assumption, although the model may still be adequate for prediction purposes.

```{r}
plot_qq(residuals = residuals(lm_step), title = "QQ Plot - Linear Regression")
```

### Polynomial Regression

Polynomial regression was introduced to capture potential nonlinear relationships between the predictors and the target variable. During exploratory data analysis, scatterplots comparing the target with selected predictors revealed curved patterns, indicating that a simple linear relationship may be insufficient. By incorporating polynomial terms, the model is able to better approximate these nonlinear trends while still retaining a relatively interpretable structure compared to more complex machine learning models.

#### Train/Test Split & Folds

```{r}
split <- create_train_test_split(timetable_transformed, seed = SEED)

train_data <- split$train
test_data  <- split$test

folds <- create_folds(train_data, v = CV_N_FOLDS, seed = SEED)
```

#### Feature Selection

The same feature selection strategy was applied in this case. The process began with the full set of predictors, followed by backward elimination to reduce the feature space. Subsequently, the statistical output from `summary()` and the Variance Inflation Factor (VIF) were examined to identify predictors exhibiting high multicollinearity. Features with elevated VIF values were iteratively removed, and the model was re-estimated after each step.

The final set of removed predictors, identified as highly collinear, is stored in the `high_vif` vector. These features were then consistently excluded from both the training and testing datasets to ensure alignment between the two subsets.

```{r}
high_vif <- c('arrival_time_hour', 'arrival_seconds',
              'pickup_type3', 'route_type11', 'sin_time', 'cos_time',
              'day_typeweekday')

train_data <- drop_cols_report(train_data, high_vif)
test_data <- drop_cols_report(test_data, high_vif)
```


#### Model Fit

Exploratory analysis indicated that several predictors (interaction features (`distance_sin_interaction`, `distance_cos_interaction`) and spatial variables (`stop_lat`, `stop_lon`) ) exhibit clear non-linear relationships with the target variable. These patterns, observed as curved and oscillatory trends, cannot be adequately captured by a purely linear model.

To address this, second-degree polynomial transformations were introduced to capture curvature while preserving the interpretability of linear regression. The original linear terms were replaced with their polynomial representations to reduce redundancy, and the use of `poly(..., 2)` ensures orthogonality, improving numerical stability.

Although other variables, such as *segment_distance_m* and *arrival_time_hour*, also showed non-linearity, they were found to be redundant given the presence of engineered features. This resulted in rank deficiency, with corresponding coefficients appearing as `NA` in the model output. Consequently, polynomial terms were not applied to these predictors, as they did not provide additional explanatory value.

```{r}
lm_poly <- lm(
  segment_time_s ~ 
    . -
    distance_sin_interaction -
    distance_cos_interaction -
    stop_lat -
    stop_lon +
    poly(distance_sin_interaction, 2) +
    poly(distance_cos_interaction, 2) +
    poly(stop_lat, 2) +
    poly(stop_lon, 2),
  data = train_data
)

initial_vars_poly <- attr(terms(lm_poly), "term.labels")

lm_poly <- step(lm_poly, direction = "backward", trace = FALSE)

final_vars_poly <- attr(terms(lm_poly), "term.labels")

poly_final_formula <- formula(lm_poly)
poly_final_formula

summary(lm_poly)
```
#### Features Dropped

Here we list features that were dropped by Backward Elimination Approach.

```{r}
dropped_vars <- setdiff(initial_vars_poly, final_vars_poly)
dropped_vars
```

#### Variance Inflation Factor

Most predictors show **very low multicollinearity**. The only standout is the *segment_distance_m*. We kept this predictor, however, because it has the strongest relationship with target and dropping it would cause the drop in R2 of about 5 percent.

```{r}
vif(lm_poly)
```

#### Evaluation

Compared to linear baselines, the model shows a notable improvement in explanatory power, reaching an R² of 0.489, which means that nearly 49% of the variance in segment travel time is captured. This improvement suggests that incorporating non-linear transformations allows the model to better represent underlying relationships in the data. However, the relatively moderate reduction in absolute error highlights the presence of inherent variability and noise in real-world transport data, which limits further gains in prediction accuracy within the linear modeling framework.

```{r}
predict_lm <- function(model, new_data) {
  predict(model, newdata = new_data)
}

results_lm_poly <- eval_train_test(
  lm_poly,
  "Polynomial Regression",
  train_data,
  test_data,
  predict_fun = predict_lm
)

results_lm_poly
```

##### Predicted vs. Actual

The plot demonstrates a clear positive relationship between predicted and actual values, indicating that the polynomial regression model effectively captures the underlying structure of the data. Compared to the linear regression model, the fitted trend line is noticeably closer to the line of perfect prediction, suggesting improved alignment and a better ability to model non-linear patterns.

```{r}
plot_predicted_vs_actual(
  data = test_data,
  predictions = predict_lm(lm_poly, test_data),
  target = "segment_time_s",
  title = "Polynomial Regression: Predicted vs Actual"
)
```

#### Homoscedasticity

Although some heteroscedasticity remains at the extremes, the inclusion of polynomial terms reduces systematic error and improves model fit, especially for larger target values.

```{r}
plot_scale_location(lm_poly$fitted.values, lm_poly$residuals)
```

##### Normality of Resituals

The QQ plot for the polynomial regression model shows a similar overall pattern to that of the linear regression model, with residuals closely following the reference line in the central region but deviating in the tails. This suggests that while polynomial regression improves predictive accuracy, it does not fully satisfy the normality assumption of residuals and may even introduce minor deviations in regions where the linear model performed adequately.

```{r}
plot_qq(residuals = residuals(lm_poly), title = "QQ Plot - Polynomial Regression")
```

### Lasso Regression

Lasso regression was selected as an embedded feature selection method due to its ability to simultaneously perform model fitting and variable selection. By introducing an L1 regularization penalty, Lasso shrinks regression coefficients and forces some of them to become exactly zero, effectively removing less important predictors from the model. In previous linear and polynomial regression models, feature reduction was performed using a combination of backward elimination and VIF analysis, which required more manual intervention to address multicollinearity. Therefore, Lasso was included to evaluate how effectively an automated regularization-based approach can handle feature selection and correlated predictors, while producing a more compact and potentially more stable model.

#### Train/Test Split

```{r}
split <- create_train_test_split(timetable_transformed, seed = SEED)

train_data <- split$train
test_data  <- split$test
```

#### Model Fit

Prior to training the Lasso model, the dataset was transformed into a suitable format required by the glmnet package. Subsequently, 5-fold cross-validation was applied to determine the optimal regularization parameter (λ), enabling a balance between model complexity and predictive performance.

```{r}
glmnet_data <- prepare_glmnet_data(train_data, test_data, TARGET)

x_train <- glmnet_data$x_train
y_train <- glmnet_data$y_train

x_test <- glmnet_data$x_test
y_test <- glmnet_data$y_test


cv_lasso <- cv.glmnet(
  x_train,
  y_train,
  alpha = 1,        # LASSO
  nfolds = CV_N_FOLDS # splitting into training data into 5 parts
)
```

#### Retraining Model

The optimal regularization parameter (λ) was selected using the `lambda.1se` criterion obtained from cross-validation. Unlike `lambda.min`, which yields the lowest prediction error, `lambda.1se` corresponds to the largest value of λ within one standard error of the minimum error. This results in a simpler and more regularized model, often with fewer retained features and improved generalization. Using this value, the final Lasso model was trained, allowing stronger penalization of coefficients and promoting sparsity in the resulting feature set.

```{r}
# lambda controls how strongly coefficients are penalized:
# small lambda → weaker penalty → more features retained
# large lambda → stronger penalty → more coefficients shrink to zero
# lambda.min is the value that gives the lowest cross-validation error.
best_lambda_1se <- cv_lasso$lambda.1se
best_lambda_1se

# This fits the final Lasso model using the selected best lambda.
lasso_model <- glmnet(
  x_train,
  y_train,
  alpha = 1,
  lambda = best_lambda_1se
)
```

#### Model Coefficients

The coefficients of the Lasso model show that several predictors were shrunk exactly to zero (denoted by “.”), indicating that they were removed from the model. This demonstrates the inherent feature selection capability of Lasso regression. Only a subset of variables retained non-zero coefficients, forming a more compact model while preserving the most informative predictors.

```{r}

coef_lasso <- coef(lasso_model)
coef_lasso
```

#### Selected Features

Here we can list the remaining predictors that has not been dropped by lasso model.

```{r}

selected_features <- rownames(coef_lasso)[coef_lasso[,1] != 0]
selected_features <- selected_features[selected_features != "(Intercept)"]

selected_features
```

#### Evaluation

Lasso regression achieved performance comparable to linear regression with backward elimination, with slightly higher RMSE and lower R² on both training and testing sets. While it did not outperform the linear models in predictive accuracy, Lasso provided the advantage of automatic feature selection by shrinking several coefficients to zero, resulting in a simpler and more interpretable model.

```{r}
predict_glmnet <- function(model, x) {
  as.vector(predict(model, newx = x))
}

results_lasso <- eval_train_test_glmnet(
  lasso_model,
  "Lasso",
  x_train,
  y_train,
  x_test,
  y_test
)

results_lasso
```

### Elastic Net

Elastic Net regression was applied as a second embedded feature selection method. It combines the L1 penalty of Lasso, which can shrink coefficients to zero, with the L2 penalty of Ridge, which stabilizes coefficient estimates in the presence of correlated predictors. This makes Elastic Net suitable for this dataset, where several time-related and route-related variables may be correlated. The value `alpha = 0.5` was used to balance both penalties, and `lambda.1se` was selected to obtain a simpler model with stronger regularization.

#### Train/Test Split

```{r}
split <- create_train_test_split(timetable_transformed, seed = SEED)

train_data <- split$train
test_data  <- split$test
```

#### Model Fit

```{r}
glmnet_data <- prepare_glmnet_data(train_data, test_data, TARGET)

x_train <- glmnet_data$x_train
y_train <- glmnet_data$y_train

x_test <- glmnet_data$x_test
y_test <- glmnet_data$y_test

cv_elastic <- cv.glmnet(
  x_train,
  y_train,
  alpha = 0.5,
  nfolds = CV_N_FOLDS
)

best_lambda_elastic_1se <- cv_elastic$lambda.1se
# best_lambda_elastic_1se <- cv_elastic$lambda.min
best_lambda_elastic_1se
```

#### Retraining Model

```{r}
elastic_model <- glmnet(
  x_train,
  y_train,
  alpha = 0.5,
  lambda = best_lambda_elastic_1se
)

coef_elastic <- coef(elastic_model)
coef_elastic
```

#### Selected Features

Here we list the selected features by Elastic.

```{r}
selected_features_elastic <- rownames(coef_elastic)[coef_elastic[, 1] != 0]
selected_features_elastic <- selected_features_elastic[
  selected_features_elastic != "(Intercept)"
]

selected_features_elastic
length(selected_features_elastic)
```

#### Evaluation

The Elastic Net model achieved predictive performance comparable to the linear regression model with backward selection, indicating that it is able to capture the underlying relationships in the data with similar accuracy. However, a key advantage of Elastic Net lies in its ability to produce a more compact model by shrinking and eliminating less informative predictors. As a result, the final model contains fewer features, improving interpretability while reducing the risk of overfitting. In addition, the overall training process of Elastic Net is computationally more efficient compared to the iterative backward elimination procedure used in linear regression.

```{r}

predict_glmnet <- function(model, x) {
  as.vector(predict(model, newx = x))
}

results_elastic <- eval_train_test_glmnet(
  elastic_model,
  "Elastic Net",
  x_train,
  y_train,
  x_test,
  y_test
)

results_elastic
```

### Tuning & Validation Setup

Before tuning the Decision Tree and Random Forest models, a reusable validation framework was prepared. The training data were further divided into an inner training subset and a validation subset, allowing different hyperparameter combinations to be evaluated without using the final test set. This ensures that the test data remain unseen until the final model evaluation.

To make the tuning process consistent across models, generic helper functions were defined for validation-grid search and cross-validation. These functions accept model-specific fitting and prediction functions, which makes them reusable for both Decision Tree and Random Forest tuning. As a result, different hyperparameter configurations can be compared systematically using the same regression metrics, such as RMSE, MAE, and R².

#### Validation Dataset

```{r}

set.seed(SEED)

val_ids <- sample(seq_len(nrow(train_data)), size = 0.2 * nrow(train_data))

validation_data <- train_data[val_ids, ]
train_inner <- train_data[-val_ids, ]
```

#### Tuning Function

```{r}
run_validation_grid <- function(grid, fit_fun, predict_fun, train_inner, validation_data) {
  results <- list()

  for (i in seq_len(nrow(grid))) {
    params <- grid[i, ]

    model <- fit_fun(train_inner, params)

    preds <- predict_fun(model, validation_data)

    res <- evaluate_regression(
      actual = validation_data$segment_time_s,
      predicted = preds,
      model_name = paste(names(params), params, collapse = " | ")
    ) %>%
      dplyr::bind_cols(params)

    results[[i]] <- res
  }

  dplyr::bind_rows(results)
}
```

#### Cross-Validation Function

```{r}
run_cv <- function(folds, fit_fun, predict_fun, params, model_name) {
  cv_results <- purrr::map_dfr(folds$splits, function(split) {
    train_fold <- rsample::analysis(split)
    valid_fold <- rsample::assessment(split)

    model <- fit_fun(train_fold, params = params)
    preds <- predict_fun(model, valid_fold)

    evaluate_regression(
      actual = valid_fold$segment_time_s,
      predicted = preds,
      model_name = paste(model_name, "CV fold")
    )
  })

  cv_results %>%
    dplyr::summarise(
      RMSE = mean(RMSE),
      MAE  = mean(MAE),
      R2   = mean(R2)
    ) %>%
    dplyr::mutate(model = paste(model_name, "(CV)")) %>%
    dplyr::select(model, RMSE, MAE, R2)
}
```

### Decision Tree


As part of the modelling phase, Decision Trees were introduced as an alternative feature-space partitioning approach. Unlike linear and polynomial regression models, which assume global relationships between predictors and the target variable, Decision Trees recursively partition the feature space into regions with similar response values. This allows the model to naturally capture non-linear relationships and interactions without requiring explicit feature engineering.

From a modelling perspective, Decision Trees represent a balance between the interpretability of regression-based approaches (such as Linear and Polynomial Regression) and the predictive performance of more complex ensemble methods like Random Forests. While they typically achieve better flexibility and performance than simple linear models, they still maintain a relatively transparent structure that can be visualized and interpreted, making them a useful intermediate model in the overall comparison.

#### Train/Test Split & Folds

```{r}
split <- create_train_test_split(timetable_transformed, seed = SEED)

train_data <- split$train
test_data  <- split$test


folds <- create_folds(train_data, v = CV_N_FOLDS, seed = SEED)
```

#### Hyperparameter Tuning


##### Grid

The tuning grid focuses on three key hyperparameters: *maxdepth*, *minsplit*, and *cp* (complexity parameter). The *maxdepth* parameter determines the maximum depth of the tree, with values ranging from shallow (3) to relatively deep (10), allowing the model to capture both simple and more complex patterns. The *minsplit* parameter controls the minimum number of observations required to perform a split, with values of 10 and 20 chosen to prevent overly granular splits that could lead to overfitting. Finally, the *cp* parameter governs pruning by specifying the minimum improvement required for a split; smaller values (0.001) allow more complex trees, while larger values (0.01) enforce stronger regularization.

By selecting this grid our exploration ranged from shallow and more regularized trees (faster computation but risk of underfitting) to
more complex trees (require more computational time to process and also have tendency
to overfit).

```{r}
dt_grid <- expand.grid(
  maxdepth = c(3, 5, 7, 10),
  minsplit = c(10),
  cp = c(0.00001, 0.001)
)
```


minsplit was fixed to 10 after initial experiments, while maxdepth and cp were tuned because they had clearer impact on model performance.

##### Tuning

```{r}
fit_dt <- function(data, params) {
  rpart(
    segment_time_s ~ .,
    data = data,
    method = "anova",
    control = rpart.control(
      maxdepth = params$maxdepth,
      minsplit = params$minsplit,
      cp = params$cp
    )
  )
}

predict_rpart <- function(model, new_data) {
  predict(model, newdata = new_data)
}

dt_validation_results <- run_validation_grid(
  grid = dt_grid,
  fit_fun = fit_dt,
  predict_fun = predict_rpart,
  train_inner = train_inner,
  validation_data = validation_data
)

dt_validation_results
```

The results show that model performance improves consistently with increasing tree depth. As maxdepth grows from 3 to 10, RMSE and MAE decrease while R² increases (up to ~0.76), indicating that deeper trees are better able to capture the underlying structure and non-linear relationships in the data.

At the same time, varying the cp parameter has no observable impact on performance, suggesting that pruning is effectively inactive in this range and does not constrain tree growth. Overall, maxdepth is the dominant factor influencing model performance, while cp plays a negligible role under these settings.

##### Choosing Best Params

```{r}
best_dt_params <- dt_validation_results %>%
  dplyr::arrange(RMSE) %>%
  dplyr::slice(1)

best_dt_params
```

#### Cross-Validation

```{r}
cv_dt_summary <- run_cv(
  folds = folds,
  fit_fun = fit_dt,
  predict_fun = predict_rpart,
  params = best_dt_params,
  model_name = "Decision Tree"
)

cv_dt_summary
```

The cross-validation results confirm that the Decision Tree model generalizes well to unseen data. The model achieved an RMSE of approximately 20.81, MAE of 12.52, and an R² of 0.74, indicating strong predictive performance.

Compared to the validation results, the cross-validated metrics are very similar, suggesting that the model is stable and not significantly overfitting. This consistency across folds implies that the selected hyperparameters provide a good balance between model complexity and generalization ability.

Overall, the Decision Tree captures a substantial portion of the variability in the target variable while maintaining robustness across different data splits.

#### Model Fit

```{r}
dt_final <- fit_dt(train_data, best_dt_params)
```

#### Evaluation

The evaluation results indicate that the Decision Tree model performs consistently across both training and test datasets. The small gap between training and test performance suggests that the model generalizes well and does not exhibit significant overfitting.

```{r}
results_dt <- eval_train_test(dt_final, "DT", train_data, test_data, predict_fun = predict_rpart)
results_dt
```

#### Feature Importance

Here we list predictors that were most important to our Decision Tree.

```{r}
importance_dt <- dt_final$variable.importance
importance_dt_df <- data.frame(
  feature = names(importance_dt),
  importance = importance_dt
) %>%
  dplyr::arrange(desc(importance))

importance_dt_df
```

### Random Forest

To further improve predictive performance, a Random Forest model was introduced as an ensemble extension of Decision Trees. While a single Decision Tree partitions the feature space into regions, it is prone to high variance and can overfit the training data. Random Forest addresses this limitation by constructing a large number of decision trees on bootstrapped samples of the data and aggregating their predictions, typically through averaging in the regression setting.

In addition to bootstrap sampling, Random Forest introduces randomness by selecting a subset of predictors at each split, which decorrelates individual trees and improves generalization. This makes the model significantly more robust and accurate compared to a single tree, especially in the presence of complex, non-linear relationships.

Furthermore, Random Forest is robust to multicollinearity and less sensitive to outliers, while also providing an inherent measure of feature importance. These properties make it a strong candidate for comparison with simpler statistical models.

From a modelling perspective, Random Forest represents the high-performance end of the spectrum: it sacrifices interpretability compared to linear models and single decision trees, but typically achieves superior predictive accuracy.

#### Train/Test Split & Folds

```{r}
split <- create_train_test_split(timetable_transformed, seed = SEED)

train_data <- split$train
test_data  <- split$test


folds <- create_folds(train_data, v = CV_N_FOLDS, seed = SEED)
```

#### Hyperparameter tuning

To improve the performance of the Random Forest model, a simple hyperparameter tuning procedure was conducted focusing on the \`mtry\` parameter, which controls the number of predictors considered at each split. This parameter plays a crucial role in balancing model diversity and predictive strength. Two values were evaluated (mtry = 3 and mtry = 5), while keeping other parameters fixed.

The results show a substantial improvement when increasing \`mtry\` from 3 to 5. Specifically, the model with mtry = 3 achieved an R² of 0.839, whereas mtry = 5 significantly improved performance to an R² of 0.922, accompanied by a notable reduction in both RMSE and MAE. This indicates that allowing the model to consider more predictors at each split enables it to better capture relevant patterns in the data.


##### Grid

The hyperparameter grid for the Random Forest model focuses on two key parameters: *mtry* and *min.node.size*. The *mtry* parameter controls the number of predictors randomly selected at each split. This allows the model to explore different levels of randomness and feature utilization, balancing between diversity of trees and predictive strength.

The *min.node.size* parameter determines the minimum number of observations required in terminal nodes. Values were selected to compare more detailed trees (smaller nodes) with slightly more regularized ones (larger nodes), helping to assess the impact of model complexity on generalization.

```{r}
rf_grid <- expand.grid(
  mtry = c(3, 6, 9),
  min.node.size = c(5, 10)
)
```

##### Tuning

```{r}

fit_rf <- function(data, params, compute_importance = FALSE) {
  importance_type <- if (compute_importance) "impurity" else "none"

  ranger(
    segment_time_s ~ .,
    data = data,
    num.trees = 200,
    mtry = params$mtry,
    min.node.size = params$min.node.size,
    importance = importance_type,
    seed = 42
  )
}

predict_ranger <- function(model, new_data) {
  predict(model, data = new_data)$predictions
}

rf_validation_results <- run_validation_grid(
  grid = rf_grid,
  fit_fun = fit_rf,
  predict_fun = predict_ranger,
  train_inner = train_inner,
  validation_data = validation_data
)

rf_validation_results
```

The results demonstrate that the Random Forest model significantly benefits from increasing the number of predictors considered at each split (*mtry*). As *mtry* increases from 3 to 9, there is a substantial improvement in performance, with RMSE and MAE decreasing and R² increasing (up to ~0.93). This indicates that allowing the model to consider more features at each split leads to stronger and more accurate trees, likely due to the rich feature space and informative predictors in the dataset.

In contrast, the *min.node.size* parameter has a relatively minor effect on performance. Increasing it from 5 to 10 results in a slight degradation in accuracy, suggesting that smaller terminal nodes (i.e., more detailed trees) better capture the underlying structure of the data.

##### Choosing Best Params

The best-performing configuration corresponds to the model with the lowest level of regularization, characterized by a smaller *min.node.size* and a higher *mtry* value. 

This result suggests that the dataset benefits from more flexible models that can fully exploit the available feature space. By allowing deeper and more complex trees, the Random Forest is able to capture intricate non-linear relationships, leading to superior predictive performance.

```{r}
best_rf_params <- rf_validation_results %>%
  dplyr::arrange(RMSE) %>%
  dplyr::slice(1)

best_rf_params
```

#### Cross Validation

Compared to the validation results, the cross-validated metrics are very similar, suggesting that the model generalizes well and does not suffer from overfitting. The consistently high R² across folds demonstrates that the Random Forest effectively captures the complex, non-linear relationships present in the data while maintaining robustness to different data splits.

```{r}
cv_rf_summary <- run_cv(
  folds = folds,
  fit_fun = fit_rf,
  predict_fun = predict_ranger,
  params = best_rf_params,
  model_name = "Random Forest"
)

cv_rf_summary
```

#### Model Fit

The Random Forest model was implemented using the `ranger` package, which offers a more computationally efficient alternative to the standard implementation.

```{r}
rf_final <- fit_rf(train_data, params = best_rf_params, compute_importance = TRUE)
```

#### Evaluation

The Random Forest model was evaluated on both training and test datasets to assess its predictive performance and potential overfitting. Results on training data indicate an excellent fit and strong ability to capture patterns in the data. However, such high performance on training data may also suggest a tendency to overfit.

To verify generalization, the model was evaluated on unseen test data. Although there is a noticeable decrease in performance compared to the training results, the model still maintains high predictive accuracy. This gap between training and test performance indicates a moderate level of overfitting, which is expected for flexible models such as Random Forest, but does not significantly impair its ability to generalize.

```{r}
results_rf <- eval_train_test(rf_final, "RF", train_data, test_data, predict_fun = predict_ranger)
results_rf
```

#### Feature Importance

Here we list the importance of predictors used by our RF from highest to lowest.

```{r}
importance_rf <- rf_final$variable.importance
importance_rf_df <- data.frame(
  feature = names(importance_rf),
  importance = importance_rf
) %>%
  dplyr::arrange(desc(importance))

importance_rf_df
```


## Conclusion

### Model Performance Comparison

```{r}
all_results <- bind_rows(
  results_lm,
  results_lm_poly,
  results_lasso,
  results_elastic,
  results_dt,
  results_rf
)

all_results <- all_results %>%
  arrange(desc(R2)) %>%
  filter(grepl("test", model))
  
all_results
```

The comparison of test-set performance clearly highlights substantial differences between the evaluated modelling approaches. The Random Forest model achieves the best results which indicates that it is able to capture the vast majority of the variability in the target variable, reflecting its strong ability to model complex, non-linear relationships.

The Decision Tree model represents a strong intermediate solution, achieving an R² of 0.76. While its performance is significantly lower than that of Random Forest, it still substantially outperforms all regression-based models, confirming the importance of non-linear feature-space partitioning for this task.

Among the statistical models, Polynomial Regression provides the best performance, demonstrating that incorporating non-linear transformations improves predictive accuracy compared to purely linear approaches. However, the improvement remains limited relative to tree-based models. Linear Regression, Lasso, and Elastic Net exhibit very similar performance, suggesting that regularization does not provide significant benefits in this setting and that the underlying relationships cannot be adequately captured by linear assumptions alone.

#### R2

Let's compare models altogether and visualize the results using several plots. We are going to compare R², RMSE and MAE. R² is plotted separately as it uses different scale.

```{r}
ggplot(all_results, aes(x = reorder(model, R2), y = R2)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Model Comparison (Test Set)",
    x = "Model",
    y = "R²"
  )
```


#### Errors

```{r}

# Get order based on RMSE
model_order <- all_results %>%
  arrange(desc(RMSE)) %>%
  pull(model)

# Pivot + apply factor ordering
all_results_long <- all_results %>%
  pivot_longer(cols = c(RMSE, MAE), names_to = "metric", values_to = "value") %>%
  mutate(model = factor(model, levels = model_order))

# Plot
ggplot(all_results_long, aes(x = model, y = value, fill = metric)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Model Comparison Across Metrics",
    x = "Model",
    y = "Value"
  )
```
  
  

### Regression Models vs. Tree-based Models

#### Informative Features

In this section we analyze features that were used during training at least one
of the models that compose the following model families:

1. **Regression Models (LR, PR)**
2. **Tree-based Models (DT, RF)**



```{r}
extract_lm_features <- function(model, model_name) {
  terms_used <- attr(terms(model), "term.labels")
  
  tibble::tibble(
    model = model_name,
    feature = terms_used
  )
}

extract_tree_features <- function(importance_values, model_name) {
  tibble::tibble(
    model = model_name,
    feature = names(importance_values),
    importance = as.numeric(importance_values)
  ) %>%
    dplyr::arrange(desc(importance))
}
```

Was feature retained by model? This condition is expressed as boolean flag in the following table:

```{r}
# --- Extract feature sets ---
lr_features <- unique(features_lr$feature)
pr_features <- unique(features_pr$feature)

dt_features <- unique(features_dt$feature)
rf_features <- unique(features_rf$feature)

# --- Create unified feature space ---
all_features <- unique(c(
  lr_features, pr_features,
  dt_features, rf_features
))

feature_presence_reduced <- tibble(
  feature = all_features
) %>%
  mutate(
    statistical_models = feature %in% lr_features | feature %in% pr_features,
    tree_based_models  = feature %in% dt_features | feature %in% rf_features
  ) %>%
  mutate(
    priority = case_when(
      statistical_models & tree_based_models ~ 1,  # TRUE TRUE
      statistical_models | tree_based_models ~ 2,  # at least one TRUE
      TRUE ~ 3                                    # FALSE FALSE
    )
  ) %>%
  arrange(priority) %>%
  select(-priority)

feature_presence_reduced
```

The analysis reveals a substantial set of features that are consistently retained across both statistical and tree-based model families, indicating strong and robust predictive signals. These include spatial, temporal, and engineered interaction features, suggesting that the underlying phenomenon is well captured by both global and partition-based modelling approaches.

Polynomial features are exclusively utilized by statistical models, reflecting their role in explicitly modelling non-linear relationships. In contrast, tree-based models do not require such transformations, as they inherently capture non-linearity through recursive partitioning.

Conversely, several raw and highly correlated features are retained only by tree-based models. This highlights the ability of tree-based approaches to handle multicollinearity and redundancy.


### Algorithmic vs. Feature-Selection Methods

```{r}

n_features_backward <- length(coef(lm_step)) - 1
n_features_lasso <- sum(coef(lasso_model) != 0) - 1
n_features_elastic <- sum(coef(elastic_model) != 0) - 1

selection_summary <- tibble::tibble(
  model = c("Linear Regression (Backward)", "Lasso Regression", "Elastic Net"),
  # selection_type = c("Algorithmic / wrapper", "Embedded", "Embedded"),
  retained_features = c(
    n_features_backward,
    n_features_lasso,
    n_features_elastic
  )
)

selection_summary
```

#### Feature Overlap

In the table below we show whether a feature was retained in the final model by using a specific technique. This condition is demonstrated by boolean flag. Features that are by most of the methods are listed at the top.

```{r}
features_backward <- names(coef(lm_step))[-1]

features_lasso <- rownames(coef(lasso_model))[coef(lasso_model)[, 1] != 0]
features_lasso <- features_lasso[features_lasso != "(Intercept)"]

features_elastic <- rownames(coef(elastic_model))[coef(elastic_model)[, 1] != 0]
features_elastic <- features_elastic[features_elastic != "(Intercept)"]

feature_selection_table <- tibble::tibble(
  feature = unique(c(features_backward, features_lasso, features_elastic))
) %>%
  dplyr::mutate(
    backward = feature %in% features_backward,
    lasso = feature %in% features_lasso,
    elastic_net = feature %in% features_elastic,
    selected_count = backward + lasso + elastic_net
  ) %>%
  dplyr::arrange(desc(selected_count), feature)

feature_selection_table %>% select(-all_of(c('selected_count')))
```


Core subset of predictors is consistently retained across all three methods, including variables such as *segment_distance_m*, interaction features (*distance_sin_interaction*, *distance_cos_interaction*), temporal variables (*seconds_since_midnight*), and several encoded categorical features (e.g., *platform_code*, *route_*). These features can be considered highly informative and robust, as their predictive value is confirmed regardless of the selection approach.

Second, Lasso and Elastic Net retain a group of additional predictors that are excluded by backward elimination, such as *arrival_seconds*, *arrival_time_hour*, and trigonometric features (*sin_time*). These variables are likely correlated with other predictors, which leads to their removal in backward selection due to multicollinearity concerns. In contrast, regularization-based methods are more tolerant of correlated features.

Conversely, backward elimination preserves several features that are removed by both Lasso and Elastic Net, including *arrival_time_minute*, *pickup_type3*, and multiple platform-related indicators.
