---
title: "OznalProject"
output: html_document
---

# Predicting Travel Time of Bratislava MHD

## Selecting Data

For the project we selected a publicly available local open dataset available here:

<https://data.bratislava.sk/datasets/aba12fd2cbac4843bc7406151bc66106/about>

This dataset contains raw data collected by MHD Bratislava which are stored in the form of multiple text files connected by foreign keys (similar to relational databases).

The newest data update our project use was released at:

April 17, 2026

## Defining Hypothesis

The project is hypothesis-driven meaning a hypothesis will be evaluated at its end. The hypothesis is stated as follows:

*"We hypothesize that scheduled travel time between two stops in GTFS systems can be predicted from such metrics like euclidean distance, route & stop attributes, geographical metrics like Lon and Lat points, temporal features like arrival times and other metrics."*

## Selecting Scenarios

Based on the hypothesis and data structure we decided to go with these two scenarios:

1.  (Scenario 1) Compare a regression or classification setup across three statistical or machine-learning methods and two feature-space partitioning approaches. Identify which features remain informative across both method families, discuss the implications for explainability, and outline the benefits each approach offers.
2.  (Scenario 3) Extend your work by comparing one algorithmic (forward, backward, mixed selection) and two embedded feature‑selection methods (lasso, ridge, elastic net) for the same regression or classification task. This task may require some research on your side, as not all necessary information will be taught through the course. Report how many features you ultimately retain, and which features lose statistical significance as you add or remove predictors.

## Analyzing Data

As we mentioned earlier, the we data represent GTFS system of MHD Bratislava which is stored in several .txt files such as:

1.  *stops.txt* (attributes about stops located in Bratislava like Hlavná stanica, Pod stanicou, etc.)
2.  *trips.txt* (a concrete trip taken by concrete vehicle following concrete stops and of conrete route)
3.  *stop_times.txt* (attributes about stop times of concrete trip at concrete stop)
4.  routes.txt (attributes about routes such as route_type managed by MHD)
5.  and other files

Many of the files or columns are highly sparse and are not provided by MHD (e.g. *frequencies.txt*, *pathways.txt* or columns *bike_allowed*, *stop_headsign* or *shape_dist_traveled).*

When analyzing the dimensions of data it has almost **700k rows** and almost **70** which should be more than satisfactory for training simple regression models like *Linear Regression* or even more complex tree-based models like *Random Forests*.

### Target - Regression or Classification?

The hypothesis states that we are predicting travel time which looks like a typical regression problem. But after analyzing the values of our target variable we got these results:

- 0 - 165 observations

- 15 - 180 observations

- 30 - 195 observations

- ...

- 540 - 2 observations

Although the target takes **discrete values** (mostly multiples of 15 or 60 seconds), it represents an underlying **continuous physical quantity — travel time**. Therefore:

- The variable is **ordered and numeric**, not categorical

- Differences between values are meaningful (e.g., 60 vs 120 seconds)

- Discretization arises from **measurement or schedule rounding**, not from the nature of the phenomenon

As a result, **regression models are appropriate**, since they:

- preserve the **ordinal structure and magnitude differences**

- allow modeling of **continuous relationships** between predictors and travel time

- avoid loss of information that would occur if the problem were treated as classification

```         
```

## Experimental Setup

### Libraries

```{r}
library("tidyverse")
library("readr")
library("dplyr")

library("ggplot2")
library("patchwork")

library("purrr")
library("car")
library("randomForest")
library("ranger")
library("glmnet") # for lasso regression

# install.packages("glmnet")
```

### Constants

```{r}

TARGET <- 'segment_time_s'

```

## EDA

### Loading Data

Apart from loading dataset, we also perform **stratified_sampling** where we pick 10% of observations for each day_type. This ensures that each category is represented with respect to its proportion in data.

```{r}

timetable <- read_csv("../data/interim/dataset.csv")

table(timetable$day_type)

set.seed(42)

# Stratified group sample: 10% trip_id v rámci kazdeho day_type
sampled_trips <- timetable %>%
  distinct(trip_id, day_type) %>%
  group_by(day_type) %>%
  slice_sample(prop = 0.1) %>%
  ungroup() %>%
  pull(trip_id)

# Vyfiltruj vsetky riadky patriace tym tripom
timetable <- timetable %>%
  filter(trip_id %in% sampled_trips)

# Sanity check - proporcie day_type by mali byt zachovane
table(timetable$day_type)
```

Now we can take a lot at the dataset to see its structure and info about its columns.

```{r}

glimpse(timetable)
```

### Data Preparation

By performing brief initial inspection of data we noticed several problems:

1.  **Missing or sparse** (some columns clearly contain NA values)
2.  **Constant or Near-Constant Columns**
3.  **Uninformative columns** (urls, descriptions, etc.)
4.  **Unsupported columns types** (strings, time, etc.)

These problems and more will need to be dealt with during this phase. We are going to turn this unsuitable format into nice training-ready data.

#### Utility Functions

In this section we define a set of utility function that we are going to use in the later phases. Both of the functions do the same - they take a vector of columns names or row indices and drop it from the dataset. Apart from this, they also print some nice logs into console about the reduction.

```{r}
drop_cols_report <- function(data, cols_to_drop) {
  # 1. Store the initial count
  col_num_before <- ncol(data)
  
  # 2. Perform the drop
  # We use any_of() instead of all_of() to prevent errors 
  # if a column was already dropped previously.
  data_reduced <- data %>%
    select(!any_of(cols_to_drop))
  
  # 3. Calculate difference
  col_num_after <- ncol(data_reduced)
  dropped_count <- col_num_before - col_num_after
  
  # 4. Print the report
  message(paste0("Dropped cols: ", dropped_count, 
                 " - Current size: ", col_num_after))
  
  # 5. Return the cleaned dataframe
  return(data_reduced)
}

drop_rows_report <- function(data, rows_to_drop) {
  # 1. Store initial count
  row_num_before <- nrow(data)
  
  # 2. Perform the drop
  # Using -rows_to_drop handles the removal by index.
  # We check if rows_to_drop is empty first to avoid returning an empty table.
  if (length(rows_to_drop) > 0) {
    data_reduced <- data[-rows_to_drop, ]
  } else {
    data_reduced <- data
  }
  
  # 3. Calculate difference
  row_num_after <- nrow(data_reduced)
  dropped_count <- row_num_before - row_num_after
  
  # 4. Print the report
  message(paste0("Dropped rows: ", dropped_count, 
                 " - Current size: ", row_num_after))
  
  return(data_reduced)
}

find_duplicate_columns <- function(df) {
  # dup_col_indexes <- which(duplicated(as.list(df)))
  # dup_cols <- names(df)[dup_col_indexes]
  # 
  # return(dup_cols)
  is_dup <- duplicated(as.list(df))
  
  dup_summary <- tibble::tibble(
    column = names(df),
    is_duplicate = is_dup
  )
  
  return(dup_summary %>% filter(is_duplicate) %>% pull(column))
}

# =========================================================
# 1. One-hot encoding for low-cardinality predictors
#    Returns the original dataset with encoded columns added
#    and the original column removed.
# =========================================================
# one_hot_encode <- function(data, col) {
#   data[[col]] <- as.factor(data[[col]])
#   
#   mm <- model.matrix(
#     as.formula(paste0("~ `", col, "` - 1")),
#     data = data
#   ) %>%
#     as.data.frame()
#   
#   data %>%
#     bind_cols(mm) %>%
#     select(-all_of(col))
# }
one_hot_encode <- function(data, col) {
  data[[col]] <- as.factor(data[[col]])
  
  mm <- model.matrix(
    as.formula(paste0("~ `", col, "`")),
    data = data
  ) %>%
    as.data.frame()
  
  # remove intercept column
  mm <- mm %>%
    dplyr::select(-`(Intercept)`)
  
  data %>%
    dplyr::bind_cols(mm) %>%
    dplyr::select(-all_of(col))
}

# =========================================================
# 2. Frequency encoding
#    Replaces each category by its relative frequency.
#    Good for medium/high-cardinality categoricals.
# =========================================================
frequency_encode <- function(data, col, new_col = NULL) {
  if (is.null(new_col)) {
    new_col <- paste0(col, "_freq")
  }
  
  freq_tbl <- data %>%
    count(.data[[col]], name = "n") %>%
    mutate(freq = n / sum(n)) %>%
    select(all_of(col), freq)
  
  data %>%
    left_join(freq_tbl, by = col) %>%
    rename(!!new_col := freq)
}

# =========================================================
# 3. Hex color -> RGB
#    Good for route_color if you want numeric representation
#    without category explosion.
# =========================================================
hex_to_rgb_df <- function(data, col = "route_color", prefix = "route") {
  hex_to_rgb <- function(hex) {
    hex <- gsub("#", "", as.character(hex))
    if (is.na(hex) || nchar(hex) != 6) return(c(NA, NA, NA))
    c(
      strtoi(substr(hex, 1, 2), 16L),
      strtoi(substr(hex, 3, 4), 16L),
      strtoi(substr(hex, 5, 6), 16L)
    )
  }
  
  rgb_vals <- t(sapply(data[[col]], hex_to_rgb)) %>%
    as.data.frame()
  
  colnames(rgb_vals) <- c(
    paste0(prefix, "_r"),
    paste0(prefix, "_g"),
    paste0(prefix, "_b")
  )
  
  bind_cols(data, rgb_vals)
}
```

#### Handling Missing Values

Throughout the following sections we are going to do the same:

1.  Create summaries about columns or rows which are somehow problematic
2.  Extract the column names or row indexes in the form of vectors

This structured format ensures that they can be effectively dropped by the reusable functions defined earlier.

In the below code we create a summary of missing values for every columns using *is.na()* provided by base R. The summaries are then split into 2 vectors containing:

1.  **low-missing features** (0\>missing_pct\>50): can be kept and handled by methods like imputing, etc.
2.  **high-missing features**: (missing_pct\>=50): problematic - should be dropped right away

```{r}
# 1. Missing values
# =========================

missing_summary <- timetable %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "missing_count") %>%
  mutate(missing_pct = missing_count / nrow(timetable) * 100) %>%
  arrange(desc(missing_pct))

missing_summary

# ===
# 1.1 Low_missing features (can be kept - must be imputed)
# ===

low_missing <- missing_summary %>%
  filter(missing_pct > 0 & missing_pct < 50) %>%
  pull(column)

low_missing

# ===
# 1.2 High missing features (must be removed)
# ===

high_missing <- missing_summary %>%
  filter(missing_pct >= 50) %>%
  pull(column)

high_missing

```

#### Constant & Near-Constant Features

These are features that carry no or only minimal signal and would only cause model to overfit they should be dropped too.\
We identify near-constant features by calculating the ratio between the 2 largest frequencies identified in a column. If this ratio is above 20 we consider this column to be near-constant.

```{r}
# =========================
# 2. Constant & Near-Constant features
# =========================
variance_summary <- timetable %>%
  summarise(across(everything(), ~ n_distinct(.))) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "n_unique") %>%
  arrange(desc(n_unique))
variance_summary

# ===
# 2.1 Constant features (should be dropped immediately)
# ===

constant_features <- variance_summary %>%
  filter(n_unique == 1) %>% pull(column)

constant_features

# ===
# 2.2 Near-zero variance (should be dropped too, too weak predictors)
# ===


# formula: ratio = freq_1 / freq_2
nzv_summary <- timetable %>%
  summarise(across(everything(), ~ {
    freq <- sort(table(.), decreasing = TRUE)
    if (length(freq) <= 1) return(NA)
    freq[1] / freq[2]
  })) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "freq_ratio") %>%
  arrange(desc(freq_ratio))

print(nzv_summary)

# Threshold example: freq_ratio > 20
nzv_cols <- nzv_summary %>%
  filter(freq_ratio > 20) %>%
  pull(column)

print(nzv_cols)
```

#### Handling Duplicates

Now we also verify whether dataset doesn't contain duplicate rows (stored as vector of row indices) and duplicate columns (stored as vector of column names). Please note that function *find_duplicate_cols* iteratively examines column by column comparing it with previous ones. This way it also finds such columns which contain only NaN values.

```{r}

# =========================
# 4. Handling duplicates
# =========================

# ===
# 4.1 Duplicate rows (should be removed)
# ===

dup_row_indexes <- timetable %>%
  duplicated() %>%
  which()

print(paste("Duplicate rows:", dup_row_indexes))

# ===
# 4.2 Duplicate cols (should be removed)
# ===


# More efficient check for duplicate columns
# R looks at the "top level" of the data frame. duplicated() now looks at each 
# vector (column) as a single item in a list and compares them.
# dup_col_indexes <- which(duplicated(as.list(timetable)))
# dup_cols <- names(timetable)[dup_col_indexes]
dup_cols <- find_duplicate_columns(timetable)

paste("Duplicate cols")
dup_cols
```

#### High-cardinality Columns

In this section we create summary about cardinality of character columns by using *n_distinct* and *n_unique* properties. These columns will need to be handled differently from one-hot-encoding since this would result in very large dataset.

```{r}

# =========================
# 5. High cardinality categorical variables
# =========================
cat_summary <- timetable %>%
  select(where(is.character)) %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "n_unique") %>%
  arrange(desc(n_unique))

print(cat_summary)

# ===
# Columns with too many categories (characters)
# ===
high_cardinality <- cat_summary %>%
  filter(n_unique > 10)
high_cardinality

high_cardinality_cols <- high_cardinality %>% pull(column)
high_cardinality_cols
```

#### Feature Reduction

The problematic features like highly-sparse features, constant or near-constant features are quite useless for models and should be dropped right away using the utility functions.

```{r}
timetable_reduced <- drop_cols_report(timetable, high_missing)
timetable_reduced <- drop_cols_report(timetable_reduced, constant_features)
timetable_reduced <- drop_cols_report(timetable_reduced, nzv_cols)
timetable_reduced <- drop_rows_report(timetable_reduced, dup_row_indexes)
timetable_reduced <- drop_cols_report(timetable_reduced, dup_cols)

dup_cols
glimpse(timetable_reduced)

table(timetable_reduced$arrival_time_second)
# # ===
# # 7.7 Highly-correlated cols
# # ===
# 
# timetable_reduced <- drop_cols_report(timetable_reduced, highly_correlated)

# glimpse(timetable_reduced)

# ===
# 7.8 Irrelevant cols
# ===
# these were used during extraction process and were broken into several predictors
# ===

highly_correlated_char = c('arrival_time', 'service_id', 'shape_id', 'stop_id',
                           'monday', 'tuesday', 'wednesday', 'thursday', 'friday',
                           'saturday', 'sunday', 'stop_sequence')
timetable_reduced <- drop_cols_report(timetable_reduced, highly_correlated_char)

# ===
# 7.8 Irrelevant cols
# ===

# these should not be in the dataset after extraction update and remove from here!!
# irrelevant_features = c('day_type')

# timetable_reduced <- drop_cols_report(timetable_reduced, irrelevant_features)

# ===
# 7.9 Basic info after filtering
# ===
dim(timetable_reduced)
glimpse(timetable_reduced)

```

#### Handling low-missing features

Let's first analyze the low-missing features:

```{r}
missing_summary %>% filter(missing_pct < 50 & missing_pct > 0)
```

##### time_since_last_stop

This one is calculated as:

*time_since_last_stop = arrival_seconds - prev(departure_seconds)*

where *prev()* represents the previous stop within a trip (sorted by *stop_sequence*). The NaN values logically represent the first stop within a trip which doesn't have a previous stop. Thus, we apply imputation with 0:

```{r}
summary(timetable_reduced$time_since_last_stop)
```

```{r}
# ===
# 8.1 Time_since_last_stop
# ===
# These are first stops in a trip

timetable_imputed <- timetable_reduced %>%
  mutate(
    time_since_last_stop = replace_na(time_since_last_stop, 0)
  )
```

##### wheelchair_accessible

Since wheelchair_accessible mostly contains value 1 (which means accessible) and a very small portion of missing values it cannot serve as a reliable predictor and should be dropped too.

```{r}
table(timetable_imputed$wheelchair_accessible)
timetable_imputed <- drop_cols_report(timetable_imputed, c('wheelchair_accessible'))
```

```{r}
# =========================
# 9. Handling categorical values
# =========================

# # Scan all character columns and count unique values
# unique_counts <- timetable_imputed %>%
#   summarise(across(where(is.character), ~n_distinct(.))) %>%
#   pivot_longer(cols = everything(), names_to = "column_name", values_to = "unique_count")
# 
# print(unique_counts)
# print(cat_summary)
```

#### Feature Encoding

Now, the remaining preprocessing step focuses on encoding categorical and non-ordinal features. Three main techniques were applied:

1.  **One-hot encoding**: Used for categorical features with relatively low cardinality (typically fewer than 15 unique values) to preserve interpretability while avoiding excessive dimensionality.
2.  **Frequency encoding**: Applied to high-cardinality categorical features, where one-hot encoding would lead to a large number of sparse columns. This approach replaces each category with its relative frequency, retaining distributional information in a compact form.
3.  **Feature transformation and decomposition**: Certain features were transformed into more informative numerical representations. For example, hexadecimal color codes were decomposed into RGB components.

```{r}
glimpse(timetable_imputed)

timetable_encoded <- timetable_imputed %>%
  one_hot_encode("platform_code") %>%
  one_hot_encode("day_type") %>%
  one_hot_encode("zone_id") %>%
  one_hot_encode("direction_id") %>%
  # one_hot_encode("wheelchair_accessible") %>%
  one_hot_encode("pickup_type") %>%
  one_hot_encode("route_type") %>%
  hex_to_rgb_df("route_color", prefix = "route") %>%
  # frequency_encode("route_short_name") %>%
  # frequency_encode("route_id") %>%
  # frequency_encode("trip_headsign") %>%
  # frequency_encode("stop_name") %>%
  # frequency_encode("stop_code") %>%
  select(
    -route_color,
    # -route_short_name,
    # -route_id,
    # -trip_headsign,
    # -stop_name,
    # -stop_code
  )

dup_cols_2 <- find_duplicate_columns(timetable_encoded)
dup_cols_2

# timetable_encoded$daytimetable_encoded
glimpse(timetable_encoded)
```

#### Detecting duplicates

Let's check again whether there are no duplicate columns in the dataset.

```{r}
dup_cols_2 <- find_duplicate_columns(timetable_encoded)
dup_cols_2

timetable_encoded <- drop_cols_report(timetable_encoded, dup_cols_2)
```

#### Saving Dataset

```{r}
# =========================
# 10. Save dataset
# =========================
saveRDS(timetable_encoded, "./datasets/timetable_encoded.rds")
```

### Training Preparation

Now that issues such as missing values, constant or near-constant features, and appropriate encoding have been addressed, the resulting dataset represents a set of candidate predictors suitable for model training.

In this second phase of exploratory data analysis, the focus shifts from general data quality to model-specific requirements. In particular, we examine:

1.  **Multicollinearity** among predictors, which can negatively affect linear models
2.  **Extreme outliers** in feature distributions, which may influence model stability
3.  **The nature of relationships** between predictors and the target variable, including potential non-linear patterns

To support this analysis, various visualization techniques such as heatmaps, boxplots, and scatterplots are employed using libraries like ggplot2. These tools help to better understand how the data aligns with the assumptions and behavior of different regression models.

Based on these insights, appropriate transformations and feature selection strategies are proposed to improve model performance and robustness.

#### Loading Data

```{r}

timetable_encoded <- readRDS("./datasets/timetable_encoded.rds")
glimpse(timetable_encoded)

```

#### Utility Functions

This section contains utility functions that are going to be repeteadly used throughout this phase.

```{r}

select_correlated_features <- function(
    data,
    target_col,
    min_abs_cor = 0.1,
    top_n = NULL,
    always_keep = NULL
) {
  # keep only numeric columns for correlation
  numeric_data <- data %>%
    dplyr::select(where(is.numeric))
  
  # check target exists
  if (!(target_col %in% names(numeric_data))) {
    stop(paste("Target column", target_col, "must exist and be numeric."))
  }
  
  # compute correlations with target
  cor_vec <- cor(numeric_data, use = "complete.obs")[, target_col]
  
  cor_df <- data.frame(
    feature = names(cor_vec),
    correlation = as.numeric(cor_vec)
  ) %>%
    dplyr::filter(feature != target_col) %>%
    dplyr::mutate(abs_correlation = abs(correlation)) %>%
    dplyr::arrange(desc(abs_correlation))
  
  # threshold filter
  cor_df <- cor_df %>%
    dplyr::filter(abs_correlation >= min_abs_cor)
  
  # optional top_n restriction
  if (!is.null(top_n)) {
    cor_df <- cor_df %>%
      dplyr::slice_head(n = top_n)
  }
  
  # final feature list
  selected_features <- unique(c(
    target_col,
    cor_df$feature,
    always_keep
  ))
  
  # keep only existing columns (safe)
  selected_features <- intersect(selected_features, names(data))
  
  return(data %>%
           dplyr::select(all_of(selected_features)))
}

plot_correlation_heatmap <- function(df, threshold = 0.3, title = "Correlation Heatmap") {
  
  # 1. Calculate correlation matrix (only for numeric columns)
  # It's safer to filter for numeric variables automatically
  cor_matrix <- df %>%
    select(where(is.numeric)) %>%
    cor(use = "complete.obs")
  
  # 2. Convert to long format
  cor_long <- as.data.frame(as.table(cor_matrix))
  colnames(cor_long) <- c("Var1", "Var2", "Correlation")
  
  # 3. Create the plot
  ggplot(cor_long, aes(x = Var1, y = Var2, fill = Correlation)) +
    geom_tile() +
    # Dynamic label filtering based on the 'threshold' argument
    geom_text(
      aes(label = ifelse(abs(Correlation) > threshold, round(Correlation, 2), "")),
      size = 3,
      color = "black"
    ) +
    scale_fill_gradient2(
      low = "#377eb8",  # Nice blue
      mid = "white",
      high = "#e41a1c", # Nice red
      midpoint = 0,
      limit = c(-1, 1)   # Ensures the scale is always centered
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.title = element_blank(),
      panel.grid = element_blank()
    ) +
    ggtitle(title)
}

plot_hist_dist <- function(df, var_name, bins = 30, show_density = FALSE) {
  p <- ggplot(df, aes(x = !!sym(var_name)))
  
  if (show_density) {
    # If density is requested, we must map the y-axis to density
    p <- p + geom_histogram(
      aes(y = after_stat(density)),
      fill = "steelblue", 
      color = "white", 
      bins = bins, 
      alpha = 0.6
    ) +
    geom_density(color = "darkblue", linewidth = 1)
  } else {
    # Standard count-based histogram
    p <- p + geom_histogram(
      fill = "steelblue", 
      color = "white", 
      bins = bins, 
      alpha = 0.8
    )
  }
  
  p + theme_minimal() +
    labs(
      title = paste("Distribution of", var_name),
      x = var_name,
      y = if(show_density) "Density" else "Frequency"
    )
}

# plot_hist_dist <- function(df, var_name, bins = 30) {
#   ggplot(df, aes(x = !!sym(var_name))) +
#     geom_histogram(
#       fill = "steelblue", 
#       color = "white", 
#       bins = bins, 
#       alpha = 0.8
#     ) +
#     theme_minimal() +
#     labs(
#       title = paste("Distribution of", var_name),
#       x = var_name,
#       y = "Frequency"
#     )
# }

plot_box_dist <- function(df, var_name) {
  ggplot(df, aes(x = !!sym(var_name))) +
    # Adding a dummy y-axis value or a factor(0) helps boxplot orientation
    geom_boxplot(fill = "lightblue", outlier.color = "red", outlier.alpha = 0.5) +
    coord_flip() +
    theme_minimal() +
    labs(
      title = paste(var_name),
      x = NULL # Removing x label as the title covers it
    )
}

plot_predictor_boxplots <- function(
    df,
    target_col = "segment_time_s",
    ncol = 3,
    show_message = TRUE,
    title = "Outlier Detection Across Predictors",
    subtitle = "Boxplots identifying potential outliers in red"
) {
  # Identify binary columns
  is_binary <- purrr::map_lgl(df, ~ length(unique(stats::na.omit(.x))) <= 2)
  binary_cols <- names(is_binary[is_binary])
  
  # Select predictors to plot
  predictors_to_plot <- setdiff(names(df), c(target_col, binary_cols))
  
  # Optional message
  if (show_message) {
    message("Dropping binary features: ", paste(binary_cols, collapse = ", "))
  }
  
  # Create individual plots
  plot_list <- purrr::map(predictors_to_plot, ~ plot_box_dist(df, .x))
  
  # Combine plots
  combined_plot <- patchwork::wrap_plots(plot_list, ncol = ncol) +
    patchwork::plot_annotation(
      title = title,
      subtitle = subtitle,
      theme = ggplot2::theme(
        plot.title = ggplot2::element_text(size = 18, face = "bold")
      )
    )
  
  return(combined_plot)
}

```

#### Target Distribution

Now we can also analyze the distribution of our target variable

```{r}

plot_hist_dist(timetable_encoded, TARGET, bins = 30, show_density = FALSE)
# table(timetable_encoded$segment_time_s)
summary(timetable_encoded$segment_time_s)
```

Most observations are concentrated around **60 seconds**, with a gradual tail toward higher values. This indicates that **typical segment travel times are short**, while longer durations occur less frequently but extend the upper range.

#### Data Engineering

In order to better capture temporal patterns present in the data, the original time variable was transformed into a cyclical representation. Specifically, the variable seconds_since_midnight was mapped onto the unit circle using sine and cosine transformations. This approach preserves the periodic nature of time, ensuring that values close in real-world time (e.g., late night and early morning) are also close in the transformed feature space.

In addition to temporal encoding, interaction features were introduced to model the combined effect of distance and time of day. The features distance_sin_interaction and distance_cos_interaction were created by multiplying the segment distance with the corresponding sine and cosine components. These interaction terms allow the model to capture time-dependent variations in travel behavior, such as differing traffic conditions across the day.

```{r}
timetable_engineered <- add_engineered_features(timetable_encoded)

# timetable_engineered <- timetable_encoded %>%
#   mutate(
#     sin_time = sin(2 * pi * seconds_since_midnight / 86400),
#     cos_time = cos(2 * pi * seconds_since_midnight / 86400),
#     distance_sin_interaction = segment_distance_m * sin_time,
#     distance_cos_interaction = segment_distance_m * cos_time,
#   )

glimpse(timetable_engineered)
```

##### Visualizations

Now let's visualize the engineered features' distributions by plotting histograms:

```{r}
timetable_engineered %>%
  select(sin_time, cos_time, distance_sin_interaction, distance_cos_interaction) %>%
  pivot_longer(cols = everything(), names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  facet_wrap(~ feature, scales = "free") +
  theme_minimal()
```

The distributions of the cyclical features *sin_time* and *cos_time* exhibit smooth, symmetric patterns consistent with their trigonometric formulation. Both variables show higher densities near their extreme values (−1 and 1), with a gradual decline toward intermediate values, followed by a mirrored increase.

The interaction features, *distance_sin_interaction* and *distance_cos_interaction*, are centered around zero and exhibit approximately bell-shaped distributions, suggesting a near-normal structure. However, the presence of heavy tails reflects the influence of large distance values, resulting in more extreme interaction effects.

To verify the distributions of *sin_time* and *cos_time* even further we can do **circular structure check**.

```{r}
ggplot(timetable_engineered, aes(x = sin_time, y = cos_time)) +
  geom_point(alpha = 0.1) +
  coord_fixed() +
  theme_minimal()
```

The joint visualization of these features confirms their correct transformation, forming a well-defined circular structure, which indicates that the periodic nature of time has been successfully preserved.

```{r}
subset_data <- select_correlated_features(
  data = timetable_encoded,
  target_col = "segment_time_s",
  min_abs_cor = 0.05,
  always_keep = c("seconds_since_midnight", "route_short_name_freq")
)

glimpse(subset_data)


collinear_features <- c()

subset_data_filtered <- subset_data %>%
  select(-all_of(collinear_features))


# --- Engineering --- #


add_engineered_features <- function(df){
  df_engineered <- df %>%
    mutate(
      sin_time = sin(2 * pi * seconds_since_midnight / 86400),
      cos_time = cos(2 * pi * seconds_since_midnight / 86400),
      distance_sin_interaction = segment_distance_m * sin_time,
      distance_cos_interaction = segment_distance_m * cos_time,
    )
  
  return(df_engineered)
}

```

#### Checking multicollinearity & target collinearity

One of the requirements of regression models like linear or polynomial regression is that its predictors should not be heavily correlated. To verify this in a neat way we are going to print the correlation matrix in the form of **HeatMap**.

Since there is a lot of features currently in the dataset we apply this

```{r}
timetable_selected <- select_correlated_features(timetable_engineered, min_abs_cor = 0.1, target_col = "segment_time_s")

glimpse(timetable_selected)

# summary(timetable_engineered$arrival_seconds)


plot_correlation_heatmap(timetable_selected, threshold = 0.01)
```

From the observed results we can tell that there is no serious collinearity among features apart from *cos_time* and *distance_cos_interaction*

```{r}
correlated_features <- c('cos_time', 'distance_cos_interaction', 'sin_time', 'distance_sin_interaction',
                         'arrival_seconds', 'seconds_since_midnight')

plot_correlation_heatmap(
  timetable_engineered %>% 
    select(all_of(c(correlated_features, "segment_time_s")))
)

```

```{r}
highly_collineared_features <- c('zone_id101')
timetable_engineered <- drop_cols_report(timetable_engineered, highly_collineared_features)
```

#### Outliers

Extreme outliers are also problematic for regression models. Let's visualize them using boxplots.

##### Boxplots

```{r}
combined_plot <- plot_predictor_boxplots(timetable_selected)
print(combined_plot)
```

From the bopxlots that mainly distance and time feature and their interactions have a lot of outliers and that heir distributions are hard to see. Let's analyze this further by inspecting their distributions

##### Distributions

```{r}
problematic_features <- c('segment_distance_m', 'distance_cos_interaction',
                          'distance_sin_interaction', 'time_since_last_stop')

plot_boxplot <- function(data, feature){
  ggplot(data, aes(x = .data[[feature]])) +
    geom_histogram(bins = 50, fill = "steelblue") +
    # scale_x_log10() +
    theme_minimal()
}

plot_boxplot(timetable_selected, 'segment_distance_m')
plot_boxplot(timetable_selected, 'time_since_last_stop')

# summary(timetable_selected %>% select(all_of(problematic_features)))

```

##### Transformations

Based on the analysis of the predictors with most outliers and their distributions, we recommend following before using them in training:

1.  *segment_distance_m* -\> very long tail, strong right skew, **log_transform** recommended
2.  *distance_interactions* -\> also heavy tails, caused by distance metric. Log cannot be applied for negative values, so transforming distance and the recomputing is recommended 3 *time_since_last_stop* -\> only moderate skew and some extremes, recommended removal if not significant model

### Saving Data

```{r}
saveRDS(timetable_engineered, './datasets/timetable_engineered.rds')
```

## Model Training

### Utility Functions

```{r}
# ---- ALTERNATIVE (LESS VISUALLY APPEALING) ----
# plot_homoscedasticity_plot <- function(fitted, residuals){
#   data.frame(
#     fitted = fitted,
#     residuals = residuals
#   ) %>%
#     ggplot(aes(x = fitted, y = residuals)) +
#     geom_point(alpha = 0.2) +
#     geom_hline(yintercept = 0, color = "red") +
#     theme_minimal()
# }
plot_homoscedasticity_plot <- function(fitted, residuals){
  data.frame(
    fitted = fitted,
    residuals = residuals
  ) %>%
    ggplot(aes(x = fitted, y = residuals)) +
    
    # points
    geom_point(alpha = 0.1, size = 0.6, color = "steelblue") +
    
    # horizontal zero line
    geom_hline(yintercept = 0, color = "red", linewidth = 0.8) +
    
    # smooth trend (VERY IMPORTANT)
    geom_smooth(
      method = "gam",
      formula = y ~ s(x, k = 5),
      color = "orange",
      se = FALSE,
      linewidth = 1
    ) +
    
    # optional: highlight spread
    stat_density_2d(alpha = 0.2, color = "grey50") +
    
    theme_minimal() +
    labs(
      title = "Residuals vs Fitted Values",
      subtitle = "Checking homoscedasticity",
      x = "Fitted values",
      y = "Residuals"
    ) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
}

plot_scale_location <- function(fitted, residuals){
  data.frame(
    fitted = fitted,
    sqrt_abs_resid = sqrt(abs(residuals))
  ) %>%
    ggplot(aes(x = fitted, y = sqrt_abs_resid)) +
    geom_point(alpha = 0.1) +
    geom_smooth(method = "gam", se = FALSE, color = "blue") +
    theme_minimal() +
    labs(
      title = "Scale-Location Plot",
      y = "√|Residuals|"
    )
}

# --- ALTERNATIVE TO BINNED RELATIONSHIP PLOT ---
# plot_scatterplot_line <- function(data, x, y = "segment_time_s") {
#   ggplot(data, aes(x = .data[[x]], y = .data[[y]])) +
#     geom_point(alpha = 0.08, size = 0.6) +
#     geom_smooth(method = "lm", color = "red", se = FALSE, linewidth = 0.8) +
#     geom_smooth(
#       method = "gam",
#       formula = y ~ s(x, k = 5),
#       color = "blue",
#       se = FALSE,
#       linewidth = 0.9
#     ) +
#     theme_minimal() +
#     labs(
#       title = x,
#       x = x,
#       y = y
#     ) +
#     theme(
#       plot.title = element_text(size = 10, face = "bold"),
#       axis.text = element_text(size = 8),
#       axis.title = element_text(size = 8)
#     )
# }
# scatterplot_list <- purrr::map(
#   features_to_check,
#   ~ plot_scatterplot_line(sample_data, .x)
# )
# 
# combined_plot <- patchwork::wrap_plots(scatterplot_list, ncol = 2)
# 
# combined_plot

plot_binned_relationship <- function(data, x, y = "segment_time_s", bins = 30, 
                                     label_size = 12, text_size = 10) {
  data %>%
    mutate(bin = ntile(.data[[x]], bins)) %>%
    group_by(bin) %>%
    summarise(
      x_mean = mean(.data[[x]], na.rm = TRUE),
      y_mean = mean(.data[[y]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    ggplot(aes(x = x_mean, y = y_mean)) +
    geom_point(size = 1.2, alpha = 0.4) +
    geom_smooth(aes(color = "Linear"), method = "lm", se = FALSE) +
    geom_smooth(
      aes(color = "Non-linear"),
      method = "gam",
      formula = y ~ s(x, k = 5),
      se = FALSE
    ) +
    scale_color_manual(
      name = "Legend", # This name must be identical across all plots
      values = c("Linear" = "red", "Non-linear" = "blue")
    ) +
    theme_minimal() +
    theme(
      axis.title = element_text(size = label_size),
      axis.text = element_text(size = text_size)
    ) +
    labs(x = x, y = paste("Mean", y))
}

# ---- Evaluation ---- #

evaluate_regression <- function(actual, predicted, round_to = NULL, model_name = "model") {
  if (!is.null(round_to)) {
    predicted <- round(predicted / round_to) * round_to
  }
  
  rmse <- sqrt(mean((actual - predicted)^2, na.rm = TRUE))
  mae  <- mean(abs(actual - predicted), na.rm = TRUE)
  
  ss_total <- sum((actual - mean(actual, na.rm = TRUE))^2, na.rm = TRUE)
  ss_res   <- sum((actual - predicted)^2, na.rm = TRUE)
  r2 <- 1 - (ss_res / ss_total)
  
  tibble::tibble(
    model = model_name,
    RMSE = rmse,
    MAE = mae,
    R2 = r2
  )
}

plot_predicted_vs_actual <- function(
  data,
  predictions,
  target = "segment_time_s",
  title = "Predicted vs Actual Values",
  add_smooth = TRUE
) {
  plot_data <- data.frame(
    actual = data[[target]],
    predicted = predictions
  )
  
  # Create perfect line data
  line_data <- data.frame(
    x = range(plot_data$actual)
  )
  line_data$y <- line_data$x
  
  p <- ggplot(plot_data, aes(x = actual, y = predicted)) +
    
    geom_point(alpha = 0.25, size = 1, color = "steelblue") +
    
    # Perfect prediction line (now mapped → legend works)
    geom_line(
      data = line_data,
      aes(x = x, y = y, color = "Perfect prediction"),
      linewidth = 1,
      linetype = "dashed"
    )
  
  if (add_smooth) {
    p <- p + geom_smooth(
      aes(color = "Model trend"),
      method = "gam",
      formula = y ~ s(x, k = 5),
      se = FALSE,
      linewidth = 1
    )
  }
  
  p +
    scale_color_manual(
      name = "Lines",
      values = c(
        "Perfect prediction" = "red",
        "Model trend" = "darkorange"
      )
    ) +
    coord_equal() +
    theme_minimal() +
    labs(
      title = title,
      subtitle = "Comparison of predictions vs actual values",
      x = "Actual value",
      y = "Predicted value"
    ) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "bottom"
    )
}

# ---- Feature Selection ---- #

select_model_features <- function(data, features = NULL, target = "segment_time_s") {
  
  # If NULL or empty → return full dataset
  if (is.null(features) || length(features) == 0) {
    return(data)
  }
  
  keep_cols <- unique(c(target, features))
  
  missing_cols <- setdiff(keep_cols, names(data))
  if (length(missing_cols) > 0) {
    warning("Missing columns: ", paste(missing_cols, collapse = ", "))
  }
  
  data %>%
    dplyr::select(dplyr::any_of(keep_cols))
}

# ---- ENCODING ----- #

fit_frequency_encoding <- function(train_df, col) {
  freq_col <- paste0(col, "_freq")
  
  freq_table <- train_df %>%
    dplyr::count(.data[[col]])
  
  names(freq_table)[names(freq_table) == "n"] <- freq_col
  
  freq_table[[freq_col]] <- freq_table[[freq_col]] / nrow(train_df)
  
  freq_table
}

apply_frequency_encoding <- function(df, freq_table, col) {
  freq_col <- paste0(col, "_freq")
  
  df %>%
    dplyr::left_join(freq_table, by = col) %>%
    dplyr::mutate(
      !!freq_col := ifelse(is.na(.data[[freq_col]]), 0, .data[[freq_col]])
    ) %>%
    dplyr::select(-dplyr::all_of(col))
}

```

### Loading Data

The predictor `time_since_last_stop` was removed prior to model training, as it may introduce data leakage by incorporating temporal information that is not available at prediction time. However, this feature was retained during the exploratory data analysis phase, where it was frequently used for understanding relationships within the data. To avoid disrupting the consistency and completeness of the EDA process, the feature was only excluded at the modeling stage, ensuring both valid analysis and reliable model evaluation.

```{r}
timetable <- readRDS("./datasets/timetable_engineered.rds")
# glimpse(timetable)library("ranger")

glimpse(timetable)

features_to_drop <- c('time_since_last_stop')

timetable <- drop_cols_report(timetable, features_to_drop)



```

### Transformations

In EDA we identified these predictors with highly skewed distributions:

1.  *segment_distance_m*
2.  *distance_sin_interaction*
3.  *distance_cos_interaction*

We recommended applying **log transformation** only to the first feature because

1.  the last 2 feature are calculated from it (dist \* sin_time, dist \* cos_time)
2.  logarithmic transformations cannot be applied to negative values

Thus after transfromation, we only recompute the remaning two features.

```{r}
# plot_predictor_boxplots(train_data, target_col = 'segment_time_s')

# summary(train_data %>% select(all_of(c('stop_lat', 'arrival_time_second', 'segment_distance_m'))))

problematic_features <- c('segment_distance_m', 'distance_sin_interaction',
                          'distance_cos_interaction')

timetable_transformed <- timetable %>%
  mutate(
    segment_distance_m = log1p(segment_distance_m)
  )

timetable_transformed <- add_engineered_features(timetable_transformed)

plot_predictor_boxplots(timetable_transformed %>% select(problematic_features), target_col = 'segment_time_s')

# glimpse(timetable_transformed)

```

From the box plots above it is clear that the skewness has been successfully eliminated for all the problematic features.

### Checking Relationships

We are going to check the relationships between some of the most useful predictors and target. For the sake of performance, we are going to limit the total size of the dataset. For the sake of clarity, we are going to use binned scatterplots that:

1.  Split into x and y axis to a limited number of bins
2.  For each bin an average is computed
3.  For each pair a point is added

This approach notably limits the total number of points in the plot and improves overall clarity. Apart from points, the two lines are dragged through the space:

1.  linear (straight) line which indicates linear relationship
2.  non-linear (curved) line which indicates polynomial relationship

If both lines significantly overlap, it indicates linear relationship.

```{r}
sample_data <- timetable_transformed %>%
  slice_sample(n = 5000)

train_col_names <- colnames(sample_data)

train_col_names <- train_col_names[!train_col_names %in% c('segment_time_s')]

length(train_col_names)

features_to_check <- c(
  "segment_distance_m",
  "distance_cos_interaction",
  "distance_sin_interaction",
  "stop_name_freq",
  "trip_headsign_freq",
  "route_short_name_freq",
  "stop_lat",
  "stop_lon"
)

# Note: Ensure TARGET is a string, e.g., TARGET <- "my_column_name"
binned_plot_list <- purrr::map(
  features_to_check,
  ~ plot_binned_relationship(sample_data, .x, y = TARGET, bins = 30, label_size = 8, text_size = 8)
)

# guides = "collect": Patchwork looks at all the plots. If it sees that they all have a legend named "Legend" with the same colors, it removes them from the individual plots and creates one single master legend.
patchwork::wrap_plots(binned_plot_list, ncol = 2, guides = "collect") & 
  theme(legend.position = "bottom")
```

Several predictors—especially interaction terms—demonstrate clear non-linear relationships with the target, indicating that linear models alone may be insufficient and that incorporating non-linear transformations or more flexible models is justified.

### Train/Test Split

```{r}
# ===
# 12.1 Train / test split
# ===

set.seed(42)

# 1. Vyber 80% trip_id (nie riadkov)
unique_trips <- unique(timetable_transformed$trip_id)
train_trips <- sample(unique_trips, size = 0.8 * length(unique_trips))

# 2. Filter podla trip_id - cely trip ide bud do trainu, alebo do testu
train_data <- timetable_transformed %>%
  filter(trip_id %in% train_trips)

test_data <- timetable_transformed %>%
  filter(!trip_id %in% train_trips)

# 3. Sanity check - ziadny trip nesmie byt v oboch
shared_trips <- intersect(unique(train_data$trip_id), unique(test_data$trip_id))
stopifnot(length(shared_trips) == 0)

cat("Train trips:", length(unique(train_data$trip_id)), 
    "| Test trips:", length(unique(test_data$trip_id)), "\n")
cat("Train rows:", nrow(train_data), 
    "| Test rows:", nrow(test_data), "\n")

# 4. Az teraz dropnut trip_id - v modeli ho nechces
train_data <- train_data %>% select(-trip_id)
test_data  <- test_data  %>% select(-trip_id)

glimpse(train_data)


```

### Frequency Encoding

```{r}
freq_cols <- c(
  "route_short_name",
  "route_id",
  "trip_headsign",
  "stop_name",
  "stop_code"
)

glimpse(train_data)

for (col in freq_cols) {
  freq_table <- fit_frequency_encoding(train_data, col)
  
  train_data <- apply_frequency_encoding(train_data, freq_table, col)
  test_data  <- apply_frequency_encoding(test_data, freq_table, col)
}

train_data$route_short_name_freq
```

### Keep-only Features (for testing)

```{r}
features_distance <- c("segment_distance_m")

train_data <- select_model_features(train_data, features_distance)
test_data  <- select_model_features(test_data, features_distance)
```

### Backward Elimination

Backward elimination was selected as the primary feature selection method due to its ability to evaluate predictors in the context of the full model. Starting with all available features ensures that interactions and combined effects between variables are considered, which is particularly important in complex datasets where individual predictors may only be informative when used together. The method iteratively removes variables based on the Akaike Information Criterion (AIC), balancing model fit and complexity, and thus helping to prevent overfitting while maintaining predictive performance.

However, backward elimination does not explicitly address multicollinearity, which can lead to unstable coefficient estimates in regression models. Therefore, an additional feature selection step based on the Variance Inflation Factor (VIF) is applied for linear and polynomial regression models. This allows the removal of highly collinear predictors, improving model interpretability and stability.

### Linear Regression

#### Correlated Features

The features listed in `high_vif_features` were identified through an iterative process. First, a regression model was fitted using backward elimination based on the AIC criterion. Subsequently, the Variance Inflation Factor (VIF) was computed to assess multicollinearity among the remaining predictors. Features exhibiting high VIF values were then removed, and the model was refitted. This process was repeated until an acceptable level of multicollinearity was achieved.

You can skip this if you wish to keep the correlated features and use only pure backward elimination approach.

```{r}
# unimportant_features <- c('seconds_since_midnight',
#                           'stop_lat',
#                           'arrival_time_second',
#                           'platform_codeH',
#                           'platform_codeJ',
#                           'trip_headsign_freq',
#                           'direction_id1',
#                           'arrival_time_minute',
#                           'platform_codeB',
#                           'platform_codeC')
# unimportant_features <- c()
# high_vif_features <- c('sin_time', 'cos_time', 'route_type11', 'day_typesunday',
#                        'arrival_time_hour', 'arrival_seconds')
# drop_features <- append(unimportant_features, high_vif_features)


high_vif_features <- c('sin_time', 'cos_time', 'route_type11',
'day_typesunday', 'arrival_time_hour', 'arrival_seconds')


train_data <- drop_cols_report(train_data, high_vif_features)
test_data <- drop_cols_report(test_data, high_vif_features)

```

#### Model Fit & Feature Selection

```{r}

# lm_model <- lm(segment_time_s ~ ., data = train_data)
# summary(lm_model)
lm_model <- lm(segment_time_s ~ ., data = train_data)

lm_model <- step(
  lm_model,
  direction = "backward",
  trace = TRUE
)

summary(lm_model)
```

#### Train Data Evaluation

```{r}
predictions_lm <- predict(lm_model, newdata = train_data)

results_lm <- evaluate_regression(
  actual = train_data$segment_time_s,
  predicted = predictions_lm,
  model_name = "Linear Regression"
)

results_lm
```

#### VIF

The Variance Inflation Factor (VIF) analysis indicates that **multicollinearity is not a significant concern** in the final linear model. The majority of predictors exhibit very low VIF values, typically close to 1, suggesting minimal linear dependence among features.

```{r}

vif(lm_model)
```

#### Homoscedasticity

The scale-location plots reveal that the linear regression model exhibits clear heteroscedasticity, particularly for higher fitted values, where the variance of residuals increases substantially. This indicates that the model fails to adequately capture the underlying structure of the data.

```{r}
plot_scale_location(lm_model$fitted.values, lm_model$residuals)
```

#### Evaluation

```         
The linear regression model achieved an RMSE of 31.68 and MAE of 23.61, indicating moderate prediction errors on average. With an R² of 0.409, the model explains approximately 41% of the variance in segment travel time, demonstrating an improvement over the distance-only baseline. This suggests that incorporating additional predictors provides meaningful information beyond distance alone. However, the model still exhibits limitations in capturing more complex, non-linear relationships present in the data, which constrains its overall predictive performance.
```

```{r}
predictions_lm <- predict(lm_model, newdata = test_data)

results_lm <- evaluate_regression(
  actual = test_data$segment_time_s,
  predicted = predictions_lm,
  model_name = "Linear Regression"
)

results_lm
```

#### Predicted vs. Actual

```{r}
plot_predicted_vs_actual(
  data = test_data,
  predictions = predictions_lm,
  target = "segment_time_s",
  title = "Linear Regression: Predicted vs Actual"
)
```

### Polynomial Regression

#### Feature Selection

Since polynomial regression does not inherently perform feature selection or regularization, an **iterative, model-driven approach** was adopted. The process began with a relatively rich set of predictors, including engineered and interaction features. The model was then repeatedly refined through successive training and evaluation cycles.

After each iteration, the statistical significance of predictors was assessed using the **t-values and corresponding p-values** obtained from the fitted model. Predictors with low statistical significance (i.e., high p-values) were considered uninformative and gradually removed from the feature set.

However, statistical significance alone can be misleading in the presence of **multicollinearity**, which inflates standard errors and destabilizes coefficient estimates. Therefore, the **Variance Inflation Factor (VIF)** was used as a complementary diagnostic tool. Predictors exhibiting high multicollinearity (VIF \> 5) were systematically eliminated to improve model stability and interpretability.

```{r}
# unimportant_features <- c('seconds_since_midnight', 'platform_codeD',
#                           'platform_codeG', 'day_typesaturday', 'day_typesunday',
#                           'platform_codeF',
#                           'platform_codeB', 'platform_codeC',
#                           'direction_id1', 'platform_codeJ', 'platform_codeI', 'route_type11',
#                           'arrival_seconds', 'arrival_time_hour',
#                           'route_short_name_freq', 'arrival_time_minute')

# vif>5, handles collinearity
# high_vif <- c('arrival_time_hour', 'arrival_seconds',
#               'pickup_type3', 'route_type11', 'sin_time', 'cos_time')
# 
# drop_cols <- append(unimportant_features, high_vif)
# # high_vif <- c()

high_vif <- c('arrival_time_hour', 'arrival_seconds',
              'pickup_type3', 'route_type11', 'sin_time', 'cos_time',
              'day_typeweekday')

train_data <- drop_cols_report(train_data, high_vif)
test_data <- drop_cols_report(test_data, high_vif)
```

```         
```

#### Model Fit

Based on the previous exploratory analysis, several predictors—most notably the interaction features (`distance_sin_interaction`, `distance_cos_interaction`) and spatial variables (`stop_lat`, `stop_lon`)—exhibited **clear non-linear relationships** with the target variable. In particular, the binned scatterplots revealed **curved and oscillatory patterns**, where the relationship could not be adequately captured by a simple linear trend.

To address this, **second-degree polynomial transformations** were introduced for these predictors. The goal was to allow the model to capture **curvature in the relationship**, while still maintaining the simplicity and interpretability of a linear regression framework. Quadratic terms are often sufficient to approximate moderate non-linearities, such as U-shaped or inverted U-shaped patterns observed in the data.

Importantly, the original (linear) versions of these predictors were replaced with their polynomial representations to avoid redundancy and reduce multicollinearity. The use of `poly(..., 2)` ensures that the resulting polynomial components are **orthogonal**, which stabilizes coefficient estimation and mitigates issues related to correlated predictors

```{r}
lm_poly <- lm(
  segment_time_s ~ 
    . -
    distance_sin_interaction -
    distance_cos_interaction -
    stop_lat -
    stop_lon +
    poly(distance_sin_interaction, 2) +
    poly(distance_cos_interaction, 2) +
    poly(stop_lat, 2) +
    poly(stop_lon, 2),
  data = train_data
)

lm_poly <- step(lm_poly, direction = "backward")
```

#### Evaluating Predictors

The polynomial regression model achieved an **R² of 0.505**, indicating that approximately **50.5% of the variance** in the target variable (*segment_time_s*) is explained by the selected predictors.

We can see that the polynomial terms now have more significant compared to their linear alternatives. This confirms that these predictors matter, but not linearly.

```{r}
summary(lm_poly)
length(coef(lm_poly)) - 1
```

#### Training Data Evaluation

```{r}
predictions_lm_poly <- predict(lm_poly, newdata = train_data)


results_lm_poly <- evaluate_regression(
  actual = train_data$segment_time_s,
  predicted = predictions_lm_poly,
  model_name = "Polynomial Regression"
)

results_lm_poly
```

#### Variance Inflation Factor

Most predictors show **very low multicollinearity**. The only standout is the *segment_distance_m*. We kept this predictor, however, because it has the strongest relationship with target and dropping it would cause the drop in R2 of about 5 percent.

```{r}
vif(lm_poly)
```

#### Homoscedasticity

Although some heteroscedasticity remains at the extremes, the inclusion of polynomial terms reduces systematic error and improves model fit, especially for larger target values.

```{r}

# plot_homoscedasticity_plot(lm_poly$fitted.values, lm_poly$residuals)


plot_scale_location(lm_poly$fitted.values, lm_poly$residuals)
```

#### Evaluation

The polynomial regression model achieved an RMSE of 29.46 and MAE of 21.62, indicating that the average prediction error remains in the range of approximately 20–30 seconds. Compared to linear baselines, the model shows a notable improvement in explanatory power, reaching an R² of 0.489, which means that nearly 49% of the variance in segment travel time is captured. This improvement suggests that incorporating non-linear transformations allows the model to better represent underlying relationships in the data. However, the relatively moderate reduction in absolute error highlights the presence of inherent variability and noise in real-world transport data, which limits further gains in prediction accuracy within the linear modeling framework.

```{r}
predictions_lm_poly <- predict(lm_poly, newdata = test_data)

results_lm_poly <- evaluate_regression(
  actual = test_data$segment_time_s,
  predicted = predictions_lm_poly,
  model_name = "Polynomial Regression"
)

results_lm_poly
# predictions_lm <- round(predictions_lm / 15) * 15
```

#### Predicted vs. Actual

```{r}
plot_predicted_vs_actual(
  data = test_data,
  predictions = predictions_lm_poly,
  target = "segment_time_s",
  title = "Polynomial Regression: Predicted vs Actual"
)
```

### Lasso Regression

Lasso regression was selected as an embedded feature selection method due to its ability to simultaneously perform model fitting and variable selection. By introducing an L1 regularization penalty, Lasso shrinks regression coefficients and forces some of them to become exactly zero, effectively removing less important predictors from the model. This is particularly useful in datasets with a large number of features or potential multicollinearity, as it produces a more compact and interpretable model while helping to prevent overfitting.

#### Model Fit

Prior to training the Lasso model, the dataset was transformed into a suitable format required by the glmnet package. Subsequently, 5-fold cross-validation was applied to determine the optimal regularization parameter (λ), enabling a balance between model complexity and predictive performance.

```{r}
feature_names <- setdiff(names(train_data), TARGET)
# feature_names

x_train <- as.matrix(train_data[, feature_names])
y_train <- train_data[[TARGET]]

x_test <- as.matrix(test_data[, feature_names])
y_test <- test_data[[TARGET]]


cv_lasso <- cv.glmnet(
  x_train,
  y_train,
  alpha = 1,        # LASSO
  nfolds = 5 # splitting into training data into 5 parts
)
```

#### Retraining Model

The optimal regularization parameter (λ) was selected using the `lambda.1se` criterion obtained from cross-validation. Unlike `lambda.min`, which yields the lowest prediction error, `lambda.1se` corresponds to the largest value of λ within one standard error of the minimum error. This results in a simpler and more regularized model, often with fewer retained features and improved generalization. Using this value, the final Lasso model was trained, allowing stronger penalization of coefficients and promoting sparsity in the resulting feature set.

```{r}
# lambda controls how strongly coefficients are penalized:
# small lambda → weaker penalty → more features retained
# large lambda → stronger penalty → more coefficients shrink to zero
# lambda.min is the value that gives the lowest cross-validation error.
best_lambda_1se <- cv_lasso$lambda.1se
best_lambda_1se

# This fits the final Lasso model using the selected best lambda.
lasso_model <- glmnet(
  x_train,
  y_train,
  alpha = 1,
  lambda = best_lambda_1se
)
```

#### Model Coefficients

The coefficients of the Lasso model show that several predictors were shrunk exactly to zero (denoted by “.”), indicating that they were removed from the model. This demonstrates the inherent feature selection capability of Lasso regression. Only a subset of variables retained non-zero coefficients, forming a more compact model while preserving the most informative predictors.

```{r}

coef_lasso <- coef(lasso_model)
coef_lasso
```

#### Selected Features

Here we can list the remaining predictors that has not been dropped by lasso model.

```{r}

selected_features <- rownames(coef_lasso)[coef_lasso[,1] != 0]
selected_features

selected_features <- selected_features[selected_features != "(Intercept)"]

selected_features
```

#### Evaluation

Lasso regression achieved performance comparable to linear regression with backward elimination, with slightly higher RMSE and lower R² on both training and testing sets. While it did not outperform the linear models in predictive accuracy, Lasso provided the advantage of automatic feature selection by shrinking several coefficients to zero, resulting in a simpler and more interpretable model.

```{r}

pred_train_lasso <- predict(lasso_model, newx = x_train)
pred_test_lasso  <- predict(lasso_model, newx = x_test)

results_lasso <- dplyr::bind_rows(
  evaluate_regression(
    actual = y_train,
    predicted = pred_train_lasso,
    model_name = "Lasso train"
  ),
  evaluate_regression(
    actual = y_test,
    predicted = pred_test_lasso,
    model_name = "Lasso test"
  )
)

results_lasso
```

### Elastic Net

Elastic Net regression was applied as a second embedded feature selection method. It combines the L1 penalty of Lasso, which can shrink coefficients to zero, with the L2 penalty of Ridge, which stabilizes coefficient estimates in the presence of correlated predictors. This makes Elastic Net suitable for this dataset, where several time-related and route-related variables may be correlated. The value `alpha = 0.5` was used to balance both penalties, and `lambda.1se` was selected to obtain a simpler model with stronger regularization.

#### Model Fit

```{r}

# Elastic Net: alpha between 0 and 1
set.seed(42)

cv_elastic <- cv.glmnet(
  x_train,
  y_train,
  alpha = 0.5,
  nfolds = 5
)

# best_lambda_elastic_1se <- cv_elastic$lambda.1se
best_lambda_elastic_1se <- cv_elastic$lambda.1se
best_lambda_elastic_1se

```

#### Retraining Model

```{r}
elastic_model <- glmnet(
  x_train,
  y_train,
  alpha = 0.5,
  lambda = best_lambda_elastic_1se
)

coef_elastic <- coef(elastic_model)
coef_elastic
```

#### Selected Features

```{r}
selected_features_elastic <- rownames(coef_elastic)[coef_elastic[, 1] != 0]
selected_features_elastic <- selected_features_elastic[
  selected_features_elastic != "(Intercept)"
]

selected_features_elastic
length(selected_features_elastic)
```

```{r}

pred_train_elastic <- predict(elastic_model, newx = x_train)
pred_test_elastic  <- predict(elastic_model, newx = x_test)

results_elastic <- dplyr::bind_rows(
  evaluate_regression(
    actual = y_train,
    predicted = pred_train_elastic,
    model_name = "Elastic Net train"
  ),
  evaluate_regression(
    actual = y_test,
    predicted = pred_test_elastic,
    model_name = "Elastic Net test"
  )
)

results_elastic

```

### Random Forest

Random Forest was selected as an additional modeling approach due to its ability to capture complex non-linear relationships and interactions between predictors without requiring explicit feature engineering. Furthermore, Random Forest is robust to multicollinearity and less sensitive to outliers, while also providing an inherent measure of feature importance. These properties make it a strong candidate for comparison with simpler statistical models.

#### Hyperparameter tuning

To improve the performance of the Random Forest model, a simple hyperparameter tuning procedure was conducted focusing on the \`mtry\` parameter, which controls the number of predictors considered at each split. This parameter plays a crucial role in balancing model diversity and predictive strength. Two values were evaluated (mtry = 3 and mtry = 5), while keeping other parameters fixed.\
\
The results show a substantial improvement when increasing \`mtry\` from 3 to 5. Specifically, the model with mtry = 3 achieved an R² of 0.839, whereas mtry = 5 significantly improved performance to an R² of 0.922, accompanied by a notable reduction in both RMSE and MAE. This indicates that allowing the model to consider more predictors at each split enables it to better capture relevant patterns in the data.

##### Validation Dataset

```{r}
set.seed(42)

val_ids <- sample(seq_len(nrow(train_data)), size = 0.2 * nrow(train_data))

validation_data <- train_data[val_ids, ]
train_inner <- train_data[-val_ids, ]
```

```{r}
# results <- list()
# 
# for (m in c(3, 5)) {
#   model <- randomForest(
#     segment_time_s ~ .,
#     data = train_data,
#     ntree = 200,
#     mtry = m,
#     nodesize = 10
#   )
#   
#   preds <- predict(model, test_data)
#   
#   res <- evaluate_regression(
#     actual = test_data$segment_time_s,
#     predicted = preds,
#     model_name = paste("RF mtry =", m)
#   )
#   
#   results[[as.character(m)]] <- res
# }
# 
# dplyr::bind_rows(results)
results <- list()
models <- list()

for (m in c(3, 5)) {
  model <- ranger(
    segment_time_s ~ .,
    data = train_inner,
    num.trees = 200,
    mtry = m,
    min.node.size = 10,
    seed = 42
  )
  
  preds <- predict(model, data = validation_data)$predictions
  
  res <- evaluate_regression(
    actual = validation_data$segment_time_s,
    predicted = preds,
    model_name = paste("RF validation mtry =", m)
  )
  
  results[[as.character(m)]] <- res
  models[[as.character(m)]] <- model
}

validation_results <- dplyr::bind_rows(results)
validation_results
```

##### Choosing Best Params

```{r}
best_mtry <- validation_results %>%
  dplyr::arrange(RMSE) %>%
  dplyr::slice(1) %>%
  dplyr::pull(model)

best_mtry
```

#### Model Fit

The Random Forest model was implemented using the `ranger` package, which offers a more computationally efficient alternative to the standard implementation. The model was trained using 100 trees (`num.trees = 100`), providing a balance between predictive performance and training time. The `mtry` parameter was set to 5 based on prior tuning results.

To control model complexity and reduce overfitting, the minimum node size was set to 10 (`min.node.size = 10`), ensuring that splits are only performed when sufficient data is available. Feature importance was computed using the impurity-based measure (`importance = "impurity"`), enabling identification of the most influential predictors.

Overall, this configuration reflects a compromise between efficiency and predictive accuracy, leveraging the strengths of Random Forest while maintaining reasonable computational cost.

```{r}
rf_best <- ranger(
  segment_time_s ~ .,
  data = train_data,
  num.trees = 100,
  mtry = 5,
  min.node.size = 10,
  importance = "impurity",
  seed = 42
)

```

#### Evaluation

The Random Forest model was evaluated on both training and test datasets to assess its predictive performance and potential overfitting. On the training data, the model achieved an RMSE of 7.71 and an R² of 0.965, indicating an excellent fit and strong ability to capture patterns in the data. However, such high performance on training data may also suggest a tendency to overfit.\
To verify generalization, the model was evaluated on unseen test data, where it achieved an RMSE of 12.13 and an R² of 0.913. Although there is a noticeable decrease in performance compared to the training results, the model still maintains high predictive accuracy. This gap between training and test performance indicates a moderate level of overfitting, which is expected for flexible models such as Random Forest, but does not significantly impair its ability to generalize.

```{r}
pred_train_rf <- predict(rf_best, data = train_data)$predictions
pred_test_rf  <- predict(rf_best, data = test_data)$predictions

results_rf <- dplyr::bind_rows(
  evaluate_regression(
    actual = train_data$segment_time_s,
    predicted = pred_train_rf,
    model_name = "Random Forest train"
  ),
  evaluate_regression(
    actual = test_data$segment_time_s,
    predicted = pred_test_rf,
    model_name = "Random Forest test"
  )
)

results_rf
```

#### Feature Importance

The feature importance analysis derived from the Random Forest model highlights **segment_distance_m** as the most influential predictor by a substantial margin, indicating that travel time is primarily driven by the physical distance between stops. This is consistent with domain expectations. Among the remaining predictors, spatial features such as **stop_lon** and **stop_lat** also exhibit strong importance, suggesting that location-specific factors contribute significantly to travel time variability. Engineered interaction features (**distance_sin_interaction** and **distance_cos_interaction**) rank highly as well, confirming that incorporating temporal patterns into distance-based measures enhances predictive power. Additionally, frequency-encoded categorical variables, including **stop_name_freq**, **route_short_name_freq**, and **trip_headsign_freq**, provide meaningful signal, likely capturing route-specific and operational characteristics.

```{r}
importance <- rf_best$variable.importance

importance_df <- data.frame(
  feature = names(importance),
  importance = importance
) %>%
  dplyr::arrange(desc(importance))

importance_df

importance_df %>%
  head(15) %>%
  ggplot(aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Feature Importance (Random Forest - ranger)",
    x = "Feature",
    y = "Importance"
  )
```
